---
title: "How to deconvolute bulk RNAseq"
output: html_document
date: "2025-10-13"
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE, message=FALSE, warning=FALSE, include = F}
rm(list=ls())

```


#### Load a seurat object


```{r, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({library(Seurat)
  library(viewmastR)
  library(magrittr)
  library(viridis)
  library(ggplot2)
  library(ComplexHeatmap)})

# Load single-cell data

bmm <- readRDS("/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchCancerCenter/Furlan_Lab_2025 - General/ddata/mullighan_b/cds/BoneMarrowMap.rds")

cluster_cols <- c(
  # Stromal
  "Stromal" = "#607D8B",

  # Stem / early progenitors (blues)
  "HSC" = "#08306B",
  "MPP-MkEry" = "#08519C",
  "MPP-MyLy" = "#2171B5",
  "LMPP" = "#4292C6",
  "CLP" = "#6BAED6",
  "MLP" = "#9ECAE1",
  "MLP-II" = "#C6DBEF",

  # Cycling (neutral greys; keep distinct)
  "Cycling Progenitor" = "#616161",
  "GMP-Cycle" = "#757575",
  # "pDC Cycling" = "#8A8A8A",
  "Pre-pDC Cycling" = "#9E9E9E",
  "Pro-B Cycling" = "#707070",
  "T Proliferating" = "#4D4D4D",
  "NK Proliferating" = "#6B6B6B",

  # Erythroid (magenta/pink ramp)
  "MEP" = "#7A0177",
  "BFU-E" = "#AE017E",
  "CFU-E" = "#DD3497",
  "Pro-Erythroblast" = "#F768A1",
  "Basophilic Erythroblast" = "#FA9FB5",
  "Polychromatic Erythroblast" = "#FCC5C0",
  "Orthochromatic Erythroblast" = "#FEE0D2",

  # Megakaryocyte (violets)
  "Megakaryocyte Precursor" = "#54278F",
  "Megakaryocyte" = "#2D004B",

  # Eo/Baso/Mast precursor (amber)
  "EoBasoMast Precursor" = "#FFB300",

  # Myeloid / mono (orange -> brown)
  "Early GMP" = "#F57C00",
  "GMP-Neut" = "#F9A825",
  "GMP-Mono" = "#E65100",
  "Early ProMono" = "#8D6E63",
  "Late ProMono" = "#6D4C41",
  "CD16 Mono" = "#A1887F",
  "CD14 Mono" = "#4E342E",

  # DC / pDC (teals)
  "ASDC" = "#006064",
  "cDC1" = "#00838F",
  "cDC2" = "#00ACC1",
  "Pre-cDC" = "#4DD0E1",
  "pDC" = "#009688",
  "Pre-pDC" = "#26A69A",

  # B lineage (blues; plasma = magenta)
  "Pre-ProB" = "#1565C0",
  "Small Pre-B" = "#1976D2",
  "Large Pre-B" = "#1E88E5",
  # "Large Pre B" = "#1E88E5",   # alias if your labels have a space
  "Pro-B VDJ" = "#0D47A1",
  "Immature B" = "#42A5F5",
  "Mature B" = "#90CAF9",
  "Plasma Cell" = "#D81B60",

  # T lineage (greens; CD4 deeper, CD8 more yellow-green)
  "CD4 Naive" = "#1B5E20",
  "CD4 Central Memory" = "#2E7D32",
  "CD4 Effector Memory" = "#388E3C",
  "CD4 Regulatory" = "#00C853",
  "CD8 Naive" = "#33691E",
  "CD8 Central Memory" = "#558B2F",
  "CD8 Tissue Resident Memory" = "#7CB342",
  "CD8 Effector Memory 1" = "#9CCC65",
  "CD8 Effector Memory 2" = "#C5E1A5",

  # NK (lavender/purples distinct from megak violets)
  "NK" = "#8E7CC3",
  "NK CD56high" = "#5E3C99"
)

setdiff(levels(factor(bmm$CellType)), names(cluster_cols))
setdiff(names(cluster_cols), levels(factor(bmm$CellType)))
# Seurat:
# DimPlot(obj, group.by = "celltype", cols = cluster_cols)

# ggplot:
# + scale_color_manual(values = cluster_cols)


g <- DimPlot(bmm, raster = F, cols = cluster_cols)
plotly::ggplotly(g)
```

Now we pseudobulk the samples to get a matrix of "signatures"

```{r, message=FALSE, warning=FALSE}
# Extract normalized counts and cell type labels
counts <- GetAssayData(bmm, layer = "data")  # log-normalized
cell_types <- bmm$CellType

# Aggregate by cell type and create signatures
signatures <- sapply(unique(cell_types), function(ct) {
  cells <- which(cell_types == ct)
  rowMeans(expm1(counts[, cells]))  # expm1 reverses log1p normalization
})
```

## Load bulk data

Next we load data released from the same publication that has bulk RNA seq from sorted populations of bone marrow cells.  These data are included in the viewmastR package.
```{r, message=FALSE, warning=FALSE}
# Load bulk counts and gene lengths
bulk_counts <- system.file("extdata", "deconvolute", "GSE74246_RNAseq_All_Counts.txt", package = "viewmastR")
bulk_counts <- read.csv(bulk_counts, sep = "\t")
rn <- bulk_counts$X_TranscriptID
bulk_counts$X_TranscriptID<- NULL
bulk_counts <- as.matrix(bulk_counts)
rownames(bulk_counts) <- rn

gene_info <- read.table(system.file("extdata", "deconvolute", "geneWeights_withSymbol.tsv", package = "viewmastR"), header = TRUE)
rownames(gene_info) <- gene_info$symbol
```


## Feature parity

Now we make sure the features are the same across all the data used for deconvolution.  Check the dimensions to amake sure they match.  We have 21 celltypes and 81 bulk samples.  Here we find 18,879 features in common across all the inputs

```{r, message=FALSE, warning=FALSE}
# Ensure same genes
common_genes <- intersect(intersect(rownames(signatures), rownames(bulk_counts)), gene_info$symbol)
signatures <- signatures[common_genes, ]
bulk_counts <- bulk_counts[common_genes, ]

# Get gene lengths
gene_lengths <- gene_info[common_genes, "length"]
gene_weights <- gene_info[common_genes, "weight"]

dim(signatures)
dim(bulk_counts)
length(gene_lengths)
```


## Deconvolute

You can read more on our implementation below, but here is a minimally working example using default parameters.

```{r}

t1 <- system.time(result_gd <- deconvolve_bulk(
  signatures = signatures,
  bulk_counts = bulk_counts,
  gene_lengths = gene_lengths,
  gene_weights = gene_weights, 
  max_iter = 2000,
  backend = "ndarray"
))

t2 <- system.time(result_gd <- deconvolve_bulk(
  signatures = signatures,
  bulk_counts = bulk_counts,
  gene_lengths = gene_lengths,
  gene_weights = gene_weights, 
  max_iter = 2000,
  backend = "wgpu"
))

result_gd2 <- deconvolve_bulk(
  signatures = signatures,
  bulk_counts = bulk_counts,
  gene_lengths = gene_lengths,
  gene_weights = gene_weights, 
  max_iter = 100000,
  backend = ifelse(as.logical(t1[3]>t2[3]), "wgpu", "ndarray")
)

system.time(result_em <- deconvolve_bulk(
  signatures = signatures,
  bulk_counts = bulk_counts,
  gene_lengths = gene_lengths,
  gene_weights = gene_weights, 
  max_iter = 3000,
  method = "em"
))
```

## Visualize

We will remove some of the bulk data that contains tumor cells or recombinant HSC for simplicity.  We will also plot the Intercept as column annotation so one can get a sense of the degree of signal that is unaccounted for.

```{r, message=FALSE, warning=FALSE}

data <- as.matrix(result_gd2$proportions_with_intercept)
data <- data[,!(grepl("LSC", colnames(data)) | grepl("rHSC", colnames(data)) | grepl("Blast", colnames(data)))]
cols <- pals::glasbey(13)
names(cols) <- unique(strsplit(colnames(data), "\\.") %>% sapply("[[", 2))
cola <- HeatmapAnnotation(bulk = strsplit(colnames(data), "\\.") %>% sapply("[[", 2), col = list(bulk=cols))
Heatmap(data[!rownames(data) %in% "Intercept",], top_annotation = cola, show_column_names = F, col = viridis(100), bottom_annotation = HeatmapAnnotation(Intercept = anno_points(data[rownames(data) %in% "Intercept",])), name = "GD Result")


data <- as.matrix(result_em$proportions_with_intercept)
data <- data[,!(grepl("LSC", colnames(data)) | grepl("rHSC", colnames(data)) | grepl("Blast", colnames(data)))]

cols <- pals::glasbey(13)
names(cols) <- unique(strsplit(colnames(data), "\\.") %>% sapply("[[", 2))
cola <- HeatmapAnnotation(bulk = strsplit(colnames(data), "\\.") %>% sapply("[[", 2), col = list(bulk=cols))
Heatmap(data[!rownames(data) %in% "Intercept",], top_annotation = cola, show_column_names = F, col = viridis(100), bottom_annotation = HeatmapAnnotation(Intercept = anno_points(data[rownames(data) %in% "Intercept",])), name = "EM Result")


```

```{r, message=FALSE, warning=FALSE}
df <- data.frame(GradDescent=as.vector(as.matrix(result_gd2$proportions_with_intercept)), ExpMax=as.vector(result_em$proportions_with_intercept), Celltype = rep(rownames(as.matrix(result_gd2$proportions_with_intercept)), ncol(as.matrix(result_gd2$proportions_with_intercept))), Sample = rep(strsplit(colnames(as.matrix(result_em$proportions_with_intercept)), "\\.") %>% sapply("[[", 2), each = nrow(as.matrix(result_em$proportions_with_intercept))))
g <- ggplot(df, aes(x=GradDescent, y=ExpMax, colour = Celltype))+geom_point()+scale_color_manual(values=c(cluster_cols, "Intercept"="black"))
plotly::ggplotly(g)

g <- ggplot(df, aes(x=GradDescent, y=ExpMax, colour = Sample))+geom_point()+scale_color_manual(values=pals::glasbey(17))
plotly::ggplotly(g)


cor(df$GradDescent, df$ExpMax)



hex_plot <- function(result, bulk_counts){
  cor(as.vector(result$pred_counts[gene_weights>0,]), as.vector(bulk_counts[gene_weights>0,]))
  obs_flat <- as.vector(bulk_counts[gene_weights>0,])
  pred_flat <- as.vector(result$pred_counts[gene_weights>0,])
  # Remove zeros for log scale
  keep_idx <- obs_flat > 0 & pred_flat > 0
  fit_df <- data.frame(
    Observed = log1p(obs_flat[keep_idx]),
    Predicted = log1p(pred_flat[keep_idx])
  )
  rsq <- cor(fit_df$Observed, fit_df$Predicted)^2
ggplot(fit_df, aes(x=Observed, y=Predicted))+geom_hex(bins = 100)+scale_fill_viridis_c(trans = "log10")+ggtitle(rsq)
}

```


```{r}
suppressPackageStartupMessages({
  library(Matrix)
})

# --- Benchmark: simulate pseudo-bulk from bmm and compare EM vs GD ---

# Choose simulation layer:
# - "counts" = raw UMI counts (more realistic)
# - "data"   = Seurat normalized/log1p data (linear after expm1; usually "easier")
use_layer <- "data"   # <- change to "data" if you want

expr <- GetAssayData(bmm, layer = use_layer)
expr <- expr[common_genes, , drop = FALSE]

if (use_layer == "data") {
  expr <- expm1(expr)  # undo log1p -> non-negative normalized values
}

cell_types <- factor(bmm$CellType)
ct_levels  <- levels(cell_types)

# 1) Split cells into TRAIN (signatures) and TEST (pseudo-bulk) to avoid leakage
set.seed(1)
train_frac <- 0.7

train_cells <- unlist(lapply(ct_levels, function(ct) {
  idx <- which(cell_types == ct)
  sample(idx, size = max(1, floor(length(idx) * train_frac)))
}))
test_cells <- setdiff(seq_len(ncol(expr)), train_cells)

expr_train <- expr[, train_cells, drop = FALSE]
expr_test  <- expr[, test_cells,  drop = FALSE]
ct_train   <- cell_types[train_cells]
ct_test    <- cell_types[test_cells]


# 2) Build signatures from TRAIN cells (genes x celltype)
signatures_sim <- sapply(ct_levels, function(ct) {
  cells <- which(ct_train == ct)
  Matrix::rowMeans(expr_train[, cells, drop = FALSE])
})
colnames(signatures_sim) <- ct_levels

common_genes <- intersect(rownames(signatures_sim), gene_info$symbol)

signatures_sim <- signatures_sim[common_genes, ]
gene_lengths_sim <- gene_info[common_genes, "length"]
gene_weights_sim <- gene_info[common_genes, "weight"]
names(gene_lengths_sim)<-common_genes
names(gene_weights_sim)<-common_genes


stopifnot(identical(rownames(signatures_sim), names(gene_lengths_sim)))
stopifnot(identical(rownames(signatures_sim), names(gene_weights_sim)))

# 3) Helpers: Dirichlet + pseudo-bulk simulator
rdirichlet1 <- function(alpha) {
  x <- rgamma(length(alpha), shape = alpha, rate = 1)
  x / sum(x)
}

simulate_pseudobulk <- function(expr_mat, cell_types_vec,
                               n_sims = 60,
                               n_cells = 20000,
                               alpha = 1,
                               depth = NULL,            # e.g., 2e7 to Poisson-resample to 20M reads; NULL = no resample
                               seed = 2,
                               replace_if_needed = TRUE) {
  set.seed(seed)
  ct_levels <- levels(factor(cell_types_vec))

  pools <- lapply(ct_levels, function(ct) which(cell_types_vec == ct))
  names(pools) <- ct_levels

  bulk_list <- vector("list", n_sims)
  truth_n   <- matrix(0, nrow = length(ct_levels), ncol = n_sims,
                      dimnames = list(ct_levels, paste0("sim", seq_len(n_sims))))
  truth_p   <- truth_n

  for (i in seq_len(n_sims)) {
    p <- rdirichlet1(rep(alpha, length(ct_levels)))
    n_by_ct <- as.integer(rmultinom(1, size = n_cells, prob = p))
    names(n_by_ct) <- ct_levels

    picked <- integer(0)
    for (ct in ct_levels) {
      n_ct <- n_by_ct[ct]
      if (n_ct == 0) next
      pool <- pools[[ct]]
      if (length(pool) == 0) next

      if (n_ct > length(pool)) {
        if (!replace_if_needed) n_ct <- length(pool)
        picked_ct <- sample(pool, size = n_ct, replace = replace_if_needed)
      } else {
        picked_ct <- sample(pool, size = n_ct, replace = FALSE)
      }
      picked <- c(picked, picked_ct)
    }

    bulk_vec <- Matrix::rowSums(expr_mat[, picked, drop = FALSE])

    # Optional: Poisson resample to a fixed sequencing depth (adds realistic noise)
    if (!is.null(depth)) {
      lib <- sum(bulk_vec)
      if (lib > 0) {
        lam <- as.numeric(bulk_vec) / lib * depth
        bulk_vec <- rpois(length(lam), lambda = pmax(lam, 0))
        names(bulk_vec) <- rownames(expr_mat)
      }
    }

    bulk_list[[i]] <- bulk_vec
    truth_n[, i] <- n_by_ct
    truth_p[, i] <- n_by_ct / sum(n_by_ct)
  }

  bulk_counts <- do.call(cbind, bulk_list)
  rownames(bulk_counts) <- rownames(expr_mat)
  colnames(bulk_counts) <- colnames(truth_p)

  list(bulk_counts = bulk_counts, truth_props = truth_p, truth_ncells = truth_n)
}

# 4) Simulate pseudo-bulk from TEST cells
sim <- simulate_pseudobulk(
  expr_mat        = expr_test,
  cell_types_vec  = ct_test,
  n_sims          = 80,
  n_cells         = 20000,
  alpha           = 1,     # smaller => sparser mixtures; larger => more even mixtures
  depth           = 2e7,   # set NULL if you don't want Poisson depth noise
  seed            = 2
)

bulk_counts_sim <- sim$bulk_counts
truth_props     <- sim$truth_props

# Make sure bulk genes match signatures
bulk_counts_sim <- bulk_counts_sim[rownames(signatures_sim), , drop = FALSE]

# 5) Deconvolve (GD vs EM)
t_gd <- system.time(res_gd <- deconvolve_bulk(
  signatures   = signatures_sim,
  bulk_counts  = bulk_counts_sim,
  gene_lengths = gene_lengths_sim,
  gene_weights = gene_weights_sim,
  max_iter     = 4000,
  backend      = "ndarray"
))

t_em <- system.time(res_em <- deconvolve_bulk(
  signatures   = signatures_sim,
  bulk_counts  = bulk_counts_sim,
  gene_lengths = gene_lengths_sim,
  gene_weights = gene_weights_sim,
  max_iter     = 4000,
  method       = "em"
))

# 6) Scoring
get_pred <- function(res, ct_levels) {
  P <- as.matrix(res$proportions_with_intercept)
  P <- P[rownames(P) %in% ct_levels, , drop = FALSE]  # drop intercept if present
  P <- P[ct_levels, , drop = FALSE]
  P <- pmax(P, 0)
  P <- sweep(P, 2, colSums(P) + 1e-12, "/")           # renormalize to sum-to-1
  P
}

score_preds <- function(pred, truth) {
  stopifnot(identical(dim(pred), dim(truth)))
  ns <- ncol(truth)
  out <- data.frame(sample = colnames(truth),
                    rmse = NA_real_, mae = NA_real_, tv = NA_real_, cor = NA_real_)
  for (j in seq_len(ns)) {
    p <- pred[, j]; t <- truth[, j]
    out$rmse[j] <- sqrt(mean((p - t)^2))
    out$mae[j]  <- mean(abs(p - t))
    out$tv[j]   <- 0.5 * sum(abs(p - t))              # total variation distance
    out$cor[j]  <- suppressWarnings(cor(p, t))        # may be NA if constant
  }
  out
}

pred_gd <- get_pred(res_gd, ct_levels)
pred_em <- get_pred(res_em, ct_levels)

scores_gd <- score_preds(pred_gd, truth_props); scores_gd$method <- "GD"
scores_em <- score_preds(pred_em, truth_props); scores_em$method <- "EM"
scores <- rbind(scores_gd, scores_em)

# 7) Summaries + paired comparison (lower RMSE/MAE/TV is better; higher cor is better)
summary_medians <- aggregate(cbind(rmse, mae, tv, cor) ~ method, data = scores,
                             FUN = median, na.rm = TRUE)
print(summary_medians)

# Paired tests across simulated samples
print(wilcox.test(scores_gd$rmse, scores_em$rmse, paired = TRUE))
print(wilcox.test(scores_gd$mae,  scores_em$mae,  paired = TRUE))
print(wilcox.test(scores_gd$tv,   scores_em$tv,   paired = TRUE))

# Optional: quick visualization
ggplot(scores, aes(x = method, y = rmse)) + geom_boxplot() + theme_classic()

delta <- scores_gd$rmse - scores_em$rmse   # >0 => EM better
summary(delta)
mean(delta)
median(delta)
mean(delta > 0)  # fraction of mixtures where EM beats GD

# Which mixtures favor GD?
idx_gd_better <- which(delta < 0)
length(idx_gd_better)

# Are those mixtures sparser / rarer-cell heavy?
sparsity <- apply(truth_props, 2, function(p) sum(p < 0.01))  # #types <1%
data.frame(
  sim = colnames(truth_props)[idx_gd_better],
  delta = delta[idx_gd_better],
  n_rare_lt1pct = sparsity[idx_gd_better],
  max_prop = apply(truth_props[, idx_gd_better, drop=FALSE], 2, max)
) |> head(10)

# Plot delta vs sparsity
plot(sparsity, delta, xlab = "# cell types <1%", ylab = "GD RMSE - EM RMSE")
abline(h = 0, lty = 2)
```

```{r}
data <- as.matrix(result_em$proportions_with_intercept)
data <- data[,(grepl("LSC", colnames(data)) | grepl("rHSC", colnames(data)) | grepl("Blast", colnames(data)))]

cols <- pals::glasbey(4)
names(cols) <- unique(strsplit(colnames(data), "\\.") %>% sapply("[[", 2))
cola <- HeatmapAnnotation(bulk = strsplit(colnames(data), "\\.") %>% sapply("[[", 2), col = list(bulk=cols))
Heatmap(data[!rownames(data) %in% "Intercept",], top_annotation = cola, show_column_names = F, col = viridis(100), bottom_annotation = HeatmapAnnotation(Intercept = anno_points(data[rownames(data) %in% "Intercept",])), name = "EM Result")



```

## Load leukemia data

save seurat object of leukemia count data for all

```{r}
seu <- qs2::qs_read(file.path("/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchCancerCenter/Furlan_Lab_2025 - General/ddata/leukemia_rna_atlas/cds/leukemia_atlas_seu.qs2"))
bulk_counts <- as.matrix(GetAssayData(seu, layers = "counts"))

gene_info <- read.table(system.file("extdata", "deconvolute", "geneWeights_withSymbol.tsv", package = "viewmastR"), header = TRUE)
rownames(gene_info) <- gene_info$symbol

counts <- GetAssayData(bmm, layer = "data")  # log-normalized
cell_types <- bmm$CellType

# Aggregate by cell type and create signatures
signatures <- sapply(unique(cell_types), function(ct) {
  cells <- which(cell_types == ct)
  rowMeans(expm1(counts[, cells]))  # expm1 reverses log1p normalization
})

common_genes <- intersect(intersect(rownames(signatures), rownames(bulk_counts)), gene_info$symbol)
signatures <- signatures[common_genes, ]
bulk_counts <- bulk_counts[common_genes, ]

# Get gene lengths
gene_lengths <- gene_info[common_genes, "length"]
gene_weights <- gene_info[common_genes, "weight"]

dim(signatures)
dim(bulk_counts)
length(gene_lengths)


leukemia_colors <- c(
  # AML Group 1 - APL
  "APL" = "#DC143C",                    # Crimson
  
  # AML Group 2 - Core binding factor
  "RUNX1-RUNX1T1" = "#C62828",         # Dark red
  "CBFB-MYH11" = "#722F37",            # Wine
  "CEBPAmut" = "#FFA07A",              # Light salmon
  
  # AML Group 3 - NPM1
  "NPM1mut" = "#FA8072",               # Salmon
  
  # AML Group 4 - Mixed group
  "NUP98-NSD1" = "#FF1493",            # Deep pink
  "DEK-NUP214" = "#D65076",            # Rose pink
  "KMT2A_PTD" = "#DB7093",             # Pale violet red
  "UBTF_ITD" = "#FFA500",              # Orange
  "FLT3_ITD" = "#F08080",              # Light coral
  
  # AML Group 5 - KMT2A Fusion
  "KMT2A_Fusion_AML" = "#C71585",      # Medium violet red
  
  # AML Group 6 - Fusion group
  "NUP98-KDM5A" = "#FF69B4",           # Hot pink
  "CBFA2T3-GLIS2" = "#C94C7F",         # Rose
  "RBM15-MKL1" = "#FF8C00",            # Dark orange
  
  # AML Group 7 - Other/Chromosomal
  "ETS-Fusion" = "#FF6347",            # Tomato
  "TP53mut" = "#B71C1C",               # Very dark red
  "monosomy7" = "#8B0000",             # Dark red/burgundy
  "del5q" = "#800020",                 # Burgundy
  "DNMT3Amut" = "#FFB6B6",             # Light pink
  
  # Other AML
  "Rare_Fusion" = "#FF7F50",           # Coral
  "Omics_Cold_AML" = "#FFE4E1",        # Misty rose
  "Other" = "#FFEBEE",                 # Very light pink
  
  # B-ALL subtypes (Blue-Purple family)
  "BCL2/MYC" = "#4169E1",
  "CRLF2(non-Ph-like)" = "#6495ED",
  "DUX4" = "#4682B4",
  "ETV6-RUNX1" = "#5F9EA0",
  "ETV6-RUNX1-like" = "#7AC5CD",
  "High_hyperdiploid" = "#87CEEB",
  "HLF" = "#191970",
  "iAMP21" = "#6A5ACD",
  "IKZF1_N159Y" = "#7B68EE",
  "KMT2A_ALL_like" = "#9370DB",
  "KMT2A_Fusion_ALL" = "#8A2BE2",
  "Low_hyperdiploid" = "#BA55D3",
  "Low_hypodiploid" = "#DA70D6",
  "MEF2D" = "#DDA0DD",
  "Near_haploid" = "#EE82EE",
  "PAX5_P80R" = "#8B7EC8",
  "PAX5alt" = "#9F8FBF",
  "Ph" = "#4B0082",
  "Ph-like" = "#6A0DAD",
  "TCF3-PBX1" = "#5D3FD3",
  "ZNF384" = "#7F00FF",
  "NUTM1" = "#9400D3",                 # Dark violet (moved to B-ALL)
  
  # T-ALL subtypes (Green-Teal family)
  "ZNF384-like" = "#8F00FF",
  "BCL11B" = "#2E8B57",
  "ETP-like" = "#3CB371",
  "HOXA9_TCR" = "#66CDAA",
  "KMT2A" = "#20B2AA",
  "LMO2_γδ-like" = "#48D1CC",
  "MLLT10" = "#40E0D0",
  "NKX2-1" = "#00CED1",
  "NKX2-5" = "#00BFBF",
  "NUP214" = "#5F9EA0",
  "NUP98" = "#008B8B",
  "SPI1" = "#2F4F4F",
  "STAG2&LMO2" = "#556B2F",
  "TAL1_DP-like" = "#6B8E23",
  "TAL1_αβ-like" = "#808000",
  "TLX1" = "#8FBC8F",
  "TLX3" = "#90EE90",
  "TME-enriched" = "#98FB98",
  
  # DS (Bright Yellow-Orange)
  "DS" = "#FFD700",
  
  # JMML (Amber-Brown)
  "JMML" = "#D2691E",
  
  # MPN (Chocolate Brown)
  "MPN" = "#8B4513",
  
  # TMD (Gold)
  "TMD" = "#DAA520"
)


```


## Run leukemia subsets
```{r}


st<- levels(factor(seu@meta.data$leukemia_st))


# subtype <- "NUP98-NSD1"
# subtype <- "KMT2A_Fusion_AML"
# reslist <- lapply(st, function(subtype){
#   ind<-which(seu@meta.data$leukemia_st %in% subtype)
#   if(length(ind)>1){
#   return(deconvolve_bulk(
#       signatures = signatures,
#       bulk_counts = bulk_counts[,ind],
#       gene_lengths = gene_lengths,
#       gene_weights = gene_weights, 
#       max_iter = 3000,
#       method = "em"
#     ))
#   } else {
#       return("not enough samples")
#     }
#   })

# reslist <- lapply(st, function(subtype){
#   ind<-which(seu@meta.data$leukemia_st %in% subtype)
#   deconvolve_bulk(
#       signatures = signatures,
#       bulk_counts = bulk_counts[,ind],
#       gene_lengths = gene_lengths,
#       gene_weights = gene_weights, 
#       max_iter = 3000,
#       method = "em"
#     )
# })

# names(reslist)<-st
reslist <- qs2::qs_read(file = file.path("~/Desktop/leftover_reslist.qs2"))


ind<-26
st[ind]
data <- as.matrix(reslist[[ind]]$proportions_with_intercept)
Heatmap(data[!rownames(data) %in% "Intercept",], show_column_names = F, col = viridis(100), bottom_annotation = HeatmapAnnotation(Intercept = anno_points(data[rownames(data) %in% "Intercept",])), name = "GD Result")




```

## cluster
```{r}



props <- lapply(1:length(st), function (i){
  if(length(which(seu$leukemia_subtype %in% st[i]))>1){
    return(reslist[[i]]$proportions_with_intercept)
  }
  else (return(NULL))
})

#props <- props[!unlist(lapply(props, is.null))]
props <- do.call(cbind, props)

meta <- seu@meta.data[colnames(props),]

seud <- CreateSeuratObject(counts=props, meta.data = meta)
seud <- NormalizeData(seud)
seud <- ScaleData(seud)
VariableFeatures(seud)<-rownames(seud)
seud <- RunPCA(seud)
seud <- FindNeighbors(seud)
ElbowPlot(seud, ndims = 50)
seud <- RunUMAP(seud, dims = 1:50)


g <- DimPlot(seud, group.by = "leukemia_subtype", cols = leukemia_colors, pt.size = 2)
plotly::ggplotly(g)

g <- DimPlot(seu, group.by = "leukemia_subtype", cols = leukemia_colors, pt.size = 2)
plotly::ggplotly(g)


rownames(seud)
scCustomize::FeaturePlot_scCustom(seud, features = "NK")

scCustomize::FeaturePlot_scCustom(seu, features = "NCAM1")
```



## find cell surface markers
```{r}



# qs2::qs_save(reslist, file = file.path("~/Desktop/leftover_reslist.qs2"))



surface <- lapply(1:length(st), function (i){
  if(length(which(seu$leukemia_subtype %in% st[i]))>1){
    leftovers <- extract_leftovers(reslist[[i]], signatures, bulk_counts[,seu$leukemia_subtype %in% st[i]], thresh = 0.3)
    return(filter_membrane_genes(leftovers$Gene)$kept_symbols)
  } else {
    return("not enough cases")
  }
})




surface_safe <- pbmcapply::pbmclapply(1:length(st), function (i){
  if(length(which(seu$leukemia_subtype %in% st[i]))>1){
    leftovers <- extract_leftovers(reslist[[i]], signatures, bulk_counts[,seu$leukemia_subtype %in% st[i]], thresh = 0.3)
    return(filter_membrane_genes(leftovers$Gene)$kept_symbols)
  } else {
    return("not enough cases")
  }
}, mc.cores = 16)



ind <- 24
st[ind]
gene <- surface[[ind]][3]
scCustomize::FeaturePlot_scCustom(seu, features = gene)
scCustomize::FeaturePlot_scCustom(bmm, features = gene, raster = F)



db_path <- ensure_hpa_data() 

# Read the full dataset into memory once
hpa_data <- readr::read_tsv(
  db_path, 
  col_types = readr::cols(
    `Gene name` = readr::col_character(),
    Tissue = readr::col_character(),
    nTPM = readr::col_double()
  ),
  progress = FALSE
)

# Define strict safety constraints
critical_organs <- c("heart", "lung", "liver", "kidney", "brain", "pancreas")
safety_threshold <- 1.0

surface_safe <- lapply(st, function(subtype) {
  
  # A. Validate Case Counts
  # Using matching logic compatible with vector 'st' containing subtype names
  cases <- which(seu$leukemia_subtype == subtype)
  
  if (length(cases) <= 1) {
    return(NULL) # Return NULL for consistent list structure, not a string
  }
  
  # B. Extract Candidates (Your custom function)
  # Assuming 'extract_leftovers' returns a data.frame or list with a $Gene column
  # We subset bulk_counts to the specific cases for this subtype
  leftovers <- extract_leftovers(
    reslist[[match(subtype, st)]], # Ensure index alignment
    signatures, 
    bulk_counts[, cases, drop = FALSE], 
    thresh = 0.3
  )
  
  if (is.null(leftovers) || nrow(leftovers) == 0) return(NULL)
  
  # C. Membrane Filter (Vectorized)
  # We pass ALL genes at once. Much faster.
  membrane_candidates <- filter_membrane_genes(
    leftovers$Gene, 
    mode = "surface_protein", 
    return_table = FALSE
  )
  
  if (length(membrane_candidates) == 0) return(NULL)
  
  # D. Batch Safety Check (The biggest speedup)
  # Instead of checking 1-by-1, we filter the pre-loaded HPA table
  
  # 1. Find ANY entry for these genes in critical organs > threshold
  unsafe_hits <- hpa_data %>%
    dplyr::filter(
      `Gene name` %in% membrane_candidates,
      Tissue %in% critical_organs,
      nTPM >= safety_threshold
    ) %>%
    dplyr::pull(`Gene name`) %>%
    unique()
  
  # 2. Set difference: Candidates MINUS Unsafe
  safe_targets <- setdiff(membrane_candidates, unsafe_hits)
  
  if (length(safe_targets) == 0) return(NULL)
  
  # E. Create Summary Table for this Subtype
  # Calculate a simple specificity score for the safe ones
  safe_stats <- hpa_data %>%
    dplyr::filter(`Gene name` %in% safe_targets) %>%
    dplyr::group_by(`Gene name`) %>%
    dplyr::summarise(
      max_healthy_tpm = max(nTPM),
      max_healthy_tissue = Tissue[which.max(nTPM)],
      specificity = max(nTPM) / sum(nTPM)
    ) %>%
    dplyr::arrange(max_healthy_tpm)
    
  return(safe_stats)
})

# Add names to the list for easy access later
names(surface_safe) <- st

safe <- do.call(rbind, surface_safe)
genes <- unique(safe$`Gene name`)

colMeans(seu@assays$RNA@layers$data[safe$`Gene name`,])

gidx <- 41
gene <- genes[gidx]
scCustomize::FeaturePlot_scCustom(seu, features = gene)
scCustomize::FeaturePlot_scCustom(bmm, features = gene,raster =  F)

pdf("~/Desktop/surface.pdf", height = 4, width = 5)
for(gene in genes){
  plot(scCustomize::FeaturePlot_scCustom(seu, features = gene))
}
dev.off()

scCustomize::FeaturePlot_scCustom(seu, features = "ASIC2",raster =  F)

scCustomize::FeaturePlot_scCustom(bmm, features = "COL17A1",raster =  F)

score_healthy_expression("DSC3")

```

```{r}
scCustomize::FeaturePlot_scCustom(seu, features = top_targets[31,1])
scCustomize::FeaturePlot_scCustom(bmm, features = top_targets[22,1], raster = F)




scCustomize::FeaturePlot_scCustom(seu, features = top_targets[11,1])
scCustomize::FeaturePlot_scCustom(bmm, features = top_targets[11,1], raster = F)
scCustomize::FeaturePlot_scCustom(bmm, features = "TMIGD2", raster = F)
scCustomize::FeaturePlot_scCustom(seu, features = "GPC3")


DimPlot(seu, group.by = "leukemia_subtype", cols = leukemia_colors)
scCustomize::FeaturePlot_scCustom(seu, features = "ASIC1")
scCustomize::FeaturePlot_scCustom(bmm, features = "ASIC1", raster = F)
scCustomize::FeaturePlot_scCustom(bmm, features = "PSD2", raster = F)
```


```{r, include = F, eval = F}
# in progress

# # # Extract tumor program
# tumor_prog <- extract_tumor_program(result_em, signatures, bulk_counts,
#                                      gene_names = rownames(signatures),
#                                      top_n = 500)
# # 
# # # View top tumor genes
# head(tumor_prog$tumor_upregulated, 20)
# head(tumor_prog$tumor_downregulated, 20)
# # 
# # # Visualize
# visualize_tumor_program(tumor_prog, bulk)
# # 
# # # Check consistency
#  consistency <- compare_tumor_consistency(tumor_prog)
#  length(consistency$consistent_upregulated)  # Core tumor program
# # 
# # # Export for pathway analysis
# export_tumor_genes(tumor_prog, "AML_tumor_program")

```

# Mathematical Summary of Poisson regression with gradient descent

**Inputs:** Signatures `S` (G×C), bulk counts `k` (G×N), gene lengths `L`, weights `w`  
**Output:** Exposures `E` (C×N) = cell type abundances

---

## Model

### Forward Pass
```
q = [S | 1/G] × [E; intercept]              # Molecules
ŷ[g,n] = q[g,n] × (L[g] / insert_size)     # Reads
k[g,n] ~ Poisson(ŷ[g,n])                    # Observed counts
```

### Parameterization
```
E = exp(log_E),  intercept = exp(log_intercept)
log_E[c,n] ~ Normal(-5, 0.1)  # Init breaks symmetry
```

### Loss Function
```
NLL = Σ w[g] × [k·log(k) - k - k·log(ŷ+ε) + ŷ]  # Poisson likelihood
L₁  = λ₁ × (NLL/100) × Σ E                      # Sparsity penalty
L₂  = λ₂ × (NLL/100) × Σ E²                     # Stability penalty
Loss = NLL + L₁ + L₂
```

---

## Optimization

**Two phases:**
1. **NULL:** Freeze E≈0, optimize intercept only → baseline
2. **Full:** Re-init E, optimize all parameters with Adam (lr=0.01)

**Convergence:** Stop when Δloss/loss < 10⁻⁶ AND Δsparsity < 10⁻⁴

---

## Output

```
Exposures:   E[c,n] = exp(log_E[c,n])           # Cell-equivalents
Proportions: P[c,n] = E[c,n] / Σ_c E[c,n]       # Fractions (sum to 1)
Quality:     intercept / (Σ E + intercept)       # <0.1 = good fit
```

---

## Key Parameters

| Parameter | Default | Range | Purpose |
|-----------|---------|-------|---------|
| `l1_lambda` | 0 | 0-10 | Sparsity (1=moderate, 5=strong) |
| `l2_lambda` | 0 | 0-10 | Stability (prevents extreme values) |
| `learning_rate` | 0.01 | 0.001-0.1 | Optimization step size |
| `insert_size` | 500 | 150-500 | Fragment length correction |

**Recommendation:** Start with defaults (no regularization). Add `l1_lambda=0.5-2` only if too many cell types.

---

## Properties

- **Non-negativity:** Guaranteed by exp() transform
- **Scale:** Set by signature normalization (Σ S[g,c] = 1)
- **Gene length correction:** Essential for RNA-seq (longer genes → more reads)
- **Performance:** 10-30 sec GPU, 1-5 min CPU (typical: 18k genes, 21 types, 81 samples)

# Mathematical Summary of Expectation-Maximization for Deconvolution

**Inputs:** Signatures `S` (G×C), bulk counts `k` (G×N), gene lengths `L`, weights `w`  
**Output:** Exposures `E` (C×N) = cell type abundances

---

## Probabilistic Model

### Generative Process
```
For each gene g in sample n:
  1. Each read comes from source j ∈ {cell types, noise}
     P(source=j) ∝ S[g,j] × E[j,n]
  
  2. Noise has uniform signature: S[g,noise] = 1/G
  
  3. Observed counts: k[g,n] ~ Poisson(ŷ[g,n])
     where ŷ[g,n] = Σ_j S[g,j] × E[j,n] × (L[g] / insert_size)
```

### Hidden Variable
```
z[g,j,n] = # of reads from gene g, sample n, assigned to source j
           (unobserved, inferred by EM)
```

---

## EM Algorithm

### E-step: Compute Responsibilities
```
For each source j (cell type or noise):
  contribution[g,j,n] = S[g,j] × E[j,n]
  
  responsibility[g,j,n] = contribution[g,j,n] / Σ_j' contribution[g,j',n]
  
Interpretation: "What fraction of counts for gene g came from source j?"
```

### M-step: Update Exposures
```
For each source j:
  E[j,n] = Σ_g w[g] × k[g,n] × responsibility[g,j,n]
  
Interpretation: "Total weighted counts assigned to source j"

Intercept[n] = E[noise,n]  # Learned automatically
```

### Log-Likelihood (Monitoring)
```
ŷ = [S | 1/G] × [E; intercept] × (L / insert_size)
LL = Σ w[g] × [k·log(ŷ+ε) - ŷ - k·log(k) + k]

Guarantee: LL increases every iteration (or numerical precision)
```

---

## Optimization

**Single unified phase** (no null model needed!)

```
Initialize: E[c,n] = total_counts[n] / C
            intercept[n] = 0.05 × total_counts[n]

Repeat:
  1. E-step: Compute responsibilities for all sources
  2. M-step: Update E and intercept simultaneously
  3. Check: |ΔLL| < tolerance

Convergence: Typically 100-400 iterations
```

**With L1 regularization:**
```
After M-step: E ← max(E - λ₁, 0)  # Soft-thresholding
```

---

## Output

```
Exposures:   E[c,n]                              # Cell-equivalents
Intercept:   intercept[n]                        # Noise level
Proportions: P[c,n] = E[c,n] / Σ_c E[c,n]       # Fractions (sum to 1)
Quality:     intercept / (mean(E) + intercept)   # <0.1 = good fit
```

---

## Key Parameters

| Parameter | Default | Range | Purpose |
|-----------|---------|-------|---------|
| `l1_lambda` | 0 | 0-10 | Sparsity via soft-thresholding |
| `tolerance` | 10⁻⁶ | 10⁻⁸-10⁻⁴ | Convergence criterion (ΔLL) |
| `insert_size` | 500 | 150-500 | Fragment length correction |
| `max_iterations` | 1000 | 100-10000 | Safety limit |

**Recommendation:** Defaults work well. No learning rate tuning needed!

---

## Properties

- **Non-negativity:** Guaranteed by model structure (no transforms needed)
- **Monotonic improvement:** Log-likelihood never decreases
- **No hyperparameters:** Learning rate, momentum, warmup not needed
- **Natural intercept:** Noise competes with signatures automatically
- **Scale:** Set by total counts and signature normalization
- **Gene length correction:** Essential for RNA-seq (longer genes → more reads)
- **Performance:** 15-25 sec (typical: 18k genes, 21 types, 81 samples, 300 iters)

---

## Comparison to Gradient Descent

| Aspect | EM | Gradient Descent |
|--------|-----|------------------|
| **Convergence** | Guaranteed monotonic | Can oscillate/diverge |
| **Hyperparameters** | None (just tolerance) | Learning rate, warmup |
| **Iterations** | 100-400 | 1000-3000 |
| **Null model** | Not needed | Requires separate fit |
| **Interpretation** | Missing data framework | Optimization of loss |
| **Speed** | 15-25 sec | 20-30 sec |
| **Stability** | Very stable | Needs careful tuning |

---

## Intuition

**EM treats deconvolution as:**  
"For each read, which cell type (or noise) generated it?"

- **E-step:** Probabilistically assign reads to sources
- **M-step:** Count up assignments to estimate abundances

This "soft assignment" framework is more natural than gradient-based optimization for count data, leading to stable convergence without tuning.
