---
title: "How to use viewmastR to deconvolute"
output: html_document
date: "2024-12-10"
always_allow_html: true
editor_options: 
  chunk_output_type: console
---


```{r, echo=FALSE, message=FALSE, warning=FALSE}
rm(list=ls())


if(grepl("^gizmo", Sys.info()["nodename"])){
  ROOT_DIR1<-"/fh/fast/furlan_s/experiments/MB_10X_5p/cds"
  ROOT_DIR2<-"/fh/fast/furlan_s/grp/data/ddata/BM_data"
} else {
  ROOT_DIR1<-"/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchinsonCancerCenter/Furlan_Lab - General/experiments/MB_10X_5p/cds"
  ROOT_DIR2<-"/Users/sfurlan/Fred Hutch Cancer Center/Furlan_Lab - General/datasets/fetalBM/data"
  ROOT_DIR3 <- "/Users/sfurlan/Fred Hutch Cancer Center/Furlan_Lab - General/datasets/AML/1031/coembed/data"
  ROOT_DIR4 <- "/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchinsonCancerCenter/Furlan_Lab - General/datasets/Healthy_BM_greenleaf"
}
library(Seurat)
library(scCustomize)
library(monocle3)
library(magrittr)
library(flscuts)
cols <- readRDS(file.path(ROOT_DIR4, "230329_rnaAugmented_seurat.RDS"))@misc$colors
# seuHR <- readRDS(file.path(ROOT_DIR4, "230329_rnaAugmented_seurat.RDS"))
# seuHR <- readRDS(file.path(ROOT_DIR4, "healthy_bm_greenleaf.rds"))
# seuHR <- CreateSeuratObject(counts = seuHR@assays@data$counts, meta.data = data.frame(colData(seuHR)))
# seuHR <- calculate_feature_dispersion(seuHR)
# 
# seuF <- readRDS(file.path(ROOT_DIR2, "fetal_bm_seurat.rds"))



```

#make ens to symbol
```{r}
aml <-readRDS("/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchCancerCenter/Furlan_Lab - General/datasets/AML/1031/coembed/data/AML_062822/bulk_aml_se.rds")

rowData(aml)$ens<-rowData(aml)$gene_id
rowData(aml)$ens <- strsplit(rowData(aml)$ens, "\\.") %>% sapply("[[", 1)
rowData(aml)$gene_name[duplicated(rowData(aml)$ens)]<-paste(rowData(aml)$gene_name[duplicated(rowData(aml)$ens)], "_PAR_Y")
rowData(aml)$ens[duplicated(rowData(aml)$ens)]<-rowData(aml)$gene_id[duplicated(rowData(aml)$ens)]

annot <- rowData(aml)
write.csv(annot, "/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchCancerCenter/Furlan_Lab - General/resources/annotation/GRCh38_symbol_ensembl_harmonized.csv")
```


```{r}
dat<-data.table::fread("/Users/sfurlan/Library/CloudStorage/OneDrive-FredHutchCancerCenter/computation/Analysis/ddata/MPAL/data/bulk/GSE74246_RNAseq_All_Counts.txt")
rownames(dat)<-dat$`_TranscriptID`
dat$`_TranscriptID`<-NULL
mat <- as.matrix(dat)
rownames(mat)<-rownames(dat)
```



```{r}
library(biomaRt)
library(tidyverse)
cds <- readRDS(file.path(ROOT_DIR4, "healthy_bm_greenleaf.rds"))
ensembl_list <- rownames(cds)
human <- useMart("ensembl", dataset="hsapiens_gene_ensembl")
gene_coords=getBM(attributes=c("hgnc_symbol","ensembl_gene_id", "start_position","end_position"), filters="hgnc_symbol", values=ensembl_list, mart=human)
gene_coords$size=gene_coords$end_position - gene_coords$start_position
unique(gene_coords$hgnc_symbol[duplicated(gene_coords$hgnc_symbol)])

gff<-"/Users/sfurlan/refs/GRCh38/genes.gtf"
gff <- ape::read.gff(gff, na.strings = c(".", "?"), GFF3 = F)

genes <- gff[gff$feature=="gene",]
library(purrr)
library(stringr)


attr_pattern <- '(\\w+)\\s+"([^"]*)"|(\\w+)\\s+([^;"]+)'

parse_attributes_row <- function(s) {
  m <- stringr::str_match_all(s, attr_pattern)[[1]]
  if (nrow(m) == 0) return(tibble())  # empty row guard

  keys  <- ifelse(!is.na(m[,2]), m[,2], m[,4])
  vals  <- ifelse(!is.na(m[,3]), m[,3], m[,5])

  # clean
  keys <- trimws(keys)
  vals <- trimws(vals)

  # collapse duplicates (e.g., multiple 'tag' entries)
  collapsed <- lapply(split(vals, keys), function(x) paste(unique(x), collapse = ","))

  # return a one-row tibble with named fields
  tibble(!!!collapsed)
}

# Suppose your data frame is df with a column `attributes` like the example you pasted
attrs_wide <- genes %>%
  mutate(.parsed = map(attributes, parse_attributes_row)) %>%
  unnest_wider(.parsed, names_repair = "unique") %>%  # handles any repeated keys safely
  select(-attributes)  # drop original if you like

# Inspect
glimpse(attrs_wide)


attrs_wide$width <- attrs_wide$end - attrs_wide$start

dups <- attrs_wide$gene_name[duplicated(attrs_wide$gene_name)]

final_annot <- attrs_wide[!attrs_wide$gene_name %in% dups,]
gw <- read.table("/Users/sfurlan/Fred Hutch Cancer Center/Furlan_Lab - General/datasets/fetalBM/data/geneWeights.tsv", header=T)
final_annot$weight <- gw$weight[match(final_annot$gene_id, gw$geneID)]
final_annot$weight[is.na(final_annot$weight)]<- 1.0
 
write.table(data.frame(ens=final_annot$gene_id, symbol=final_annot$gene_name, length=final_annot$width, weight = final_annot$weight), "/Users/sfurlan/Fred Hutch Cancer Center/Furlan_Lab - General/datasets/fetalBM/data/geneWeights_withSymbol.tsv", quote = F, sep = "\t")


```


```{r}

cds <- readRDS(file.path(ROOT_DIR4, "healthy_bm_greenleaf.rds"))
plot_cells(cds, color_cells_by = "BioClassification")

rownames(cds)<-rowData(cds)$gene_id
#cds <- normalized_counts(cds)
PCs <- 25 #Number of PCs for clustering
n_top <- c(3000,2500,2500) #number of features
cds<-iterative_LSI(cds, num_dim = PCs, num_features = n_top, resolution = c(2e-4, 9e-4, 1e-3), verbose = T)
cds<-reduce_dimension(cds, reduction_method = "UMAP", preprocess_method = "LSI", umap.min_dist = 0.45,  umap.n_neighbors = 35, umap.metric = "euclidean", verbose=T)
plot_cells(cds, color_cells_by = "BioClassification")


feats <- cds@int_metadata$LSI_model$features

#make a seurat object for export
counts <- cds@assays@data$counts
rownames(counts)<-rowData(cds)$gene_id
seuHR <- CreateSeuratObject(counts = counts , meta.data = data.frame(colData(cds)))
seuHR <- NormalizeData(seuHR) %>% ScaleData()
lsimat <- cds@int_metadata$LSI_model$svd$v
rownames(lsimat)<-colnames(cds)
colnames(lsimat)<-paste0("LSI_", 1:ncol(lsimat))
seuHR$lsi <- CreateDimReducObject(embeddings = lsimat, assay = "RNA")
# seuHR <- readRDS(file.path(ROOT_DIR4, "healthy_bm_greenleaf.rds"))
# seuHR <- CreateSeuratObject(counts = seuHR@assays@data$counts, meta.data = data.frame(colData(seuHR)))
# seuHR <- NormalizeData(seuHR) %>% ScaleData()
# VariableFeatures(seuHR)<- feats
# seuHR <- Signac::RunTFIDF(seuHR)
# seuHR <- Signac::RunSVD(seuHR)
VariableFeatures(seuHR)<-feats
seuHR <- RunUMAP(seuHR, dims = 1:25, reduction = "lsi", min.dist = 0.45, n.neighbors = 35, metric = "euclidean")
DimPlot(seuHR, group.by = "BioClassification")
DimPlot(seuHR, group.by = "Group")

dir.create("~/Desktop/seuHR")
flscuts::export_seurat_to_10x(seuHR, dir = "~/Desktop/seuHR")
```




```{r}

library(viewmastR)

bulk_count_data = "/Users/sfurlan/Fred Hutch Cancer Center/Furlan_Lab - General/datasets/AML/1031/coembed/data/aml_all_bulk_counts.csv"
gene_weights = "/Users/sfurlan/Fred Hutch Cancer Center/Furlan_Lab - General/datasets/fetalBM/data/geneWeights.tsv"

bulk_count_data <- read.csv(bulk_count_data)
gene_weights <- read.table(gene_weights, sep = "\t", header = T)

pseudobulk <- function(seu, by, assay = "RNA", layer = "counts"){
  meta <- seu@meta.data
  counts_mat <- GetAssayData(seu, assay = assay, layer = layer)
  # extract your cell‐type factor
  cell_types <- as.character(meta[[by]])
  unique_types <- unique(cell_types)
  
  # pre‐allocate an empty matrix: genes × n_types
  genes       <- rownames(counts_mat)
  scSigs_mat  <- matrix(0,
                        nrow = length(genes),
                        ncol = length(unique_types),
                        dimnames = list(genes, unique_types))
  
  # fill it
  for (ct in unique_types) {
    idx <- which(cell_types == ct)
    # sum counts for that cell-type
    vec <- rowSums(counts_mat[, idx, drop=FALSE])
    # normalize so they sum to 1
    scSigs_mat[, ct] <- vec / sum(vec)
  }
  scSigs_mat
}



# 1) build healthy reference from your non-malignant clusters
ref_mat <- pseudobulk(seuHR, "BioClassification" )
C_mat <- bulk_count_data$size
names(C_mat)<-bulk_count_data$X
bulk_mat <- as.matrix(bulk_count_data[,3:ncol(bulk_count_data)])
rownames(bulk_mat)<-bulk_count_data$X

common_feats <- intersect(intersect(rownames(ref_mat), names(C_mat)), gene_weights$geneID)


ref_mat <- ref_mat[common_feats, ]
C_mat <- C_mat[common_feats]
bulk_mat <- bulk_mat[common_feats,]
w_vec <- gene_weights$weight[match(common_feats, gene_weights$geneID)]


tester <- c(2116, 3432, 3391, 963, 4374, 1277, 3943, 1295, 3714, 3576, 1263, 2275, 583, 2497, 889, 1380, 4483, 4431, 3146, 1955, 1997, 133, 3573, 3622, 660, 2902, 30, 1017, 2783, 1919, 2996, 4908, 2051, 1474, 1018, 631, 180, 3758, 2229, 986, 2302, 2197, 3451, 3267, 2347, 4867, 2328, 2638, 2827, 737, 371, 4592, 647, 3349, 3562, 3482, 2389, 489, 3046, 479, 2293, 1079, 2131, 4072, 1907, 3456, 2255, 540, 4294, 3011, 4610, 2970, 4819, 3088, 3114, 584, 192, 2751, 945, 1158, 1696, 3889, 1020, 2533, 1579, 4598, 3182, 2362, 1580, 2713, 3603, 4635, 3678, 2451, 2332, 4, 2168, 3370, 3502, 305)

bulk_mat_t <- bulk_mat[,tester]
dim(bulk_mat_t)

#saveRDS(list(ref_mat, bulk_mat_t, C_mat, w_vec), file.path("~/Desktop/seuHR/test_list.RDS"))



library(viewmastR)

l<-readRDS("~/Desktop/seuHR/test_list.RDS")
ref_mat = l[[1]]
bulk_mat_t = l[[2]]
C_mat = l[[3]]
w_vec <- l[[4]]

deconvolute <- function(
    ref_mat, 
    bulk_mat, 
    length_vec, 
    weight_vec,
    init_log_exp = -5,
    lr           = 1e-3,
    l1_lambda    = 0.01,
    l2_lambda    = 0.01,
    epochs       = 300)
{
    fit <- fit_deconv(
      sigs = ref_mat,
      bulk = bulk_mat,
      mol2read = length_vec,
      w_vec = weight_vec,
      init_log_exp =init_log_exp,
      lr          = lr,
      l1_lambda   = l1_lambda,
      l2_lambda   = l2_lambda,
      epochs      = epochs)
  
        ## -- 1. sort the sample columns the way pandas .sort_index(axis=1) does ----
      loss <- fit$loss
      fit <- fit$exposures[ , order(colnames(bulk_mat))]
      # 
      # 
      # ## -- 2. split out “goodness-of-fit” rows -----------------------------------
      # gof_noms <- c("pR2", "fitCount", "obsCount")
      # gof <- fit[gof_noms, , drop = FALSE]
      # 
      # ## add log2(countRatio) = log2(fitCount / obsCount) per sample
      # gof["log2(countRatio)", ] <- log2(gof["fitCount", ] / gof["obsCount", ])
      # 
      # ## -- 3. raw exposures (everything that is *not* a GOF row) -----------------
      # expos_idx <- !rownames(fit) %in% gof_noms
      # raw <- fit[expos_idx, , drop = FALSE]
    
      ## -- 4. column-wise normalisation so each sample sums to 1 -----------------
      exposures <- sweep(fit, 2, colSums(fit), FUN = "/")
    
      ## -- 5. return a list mimicking the Python dict ----------------------------
      list(
        exposures = exposures
        # gof       = gof,
        # raw       = raw          # raw (unnormalised) exposure matrix
      )
}

# C_mat <- matrix(
#   as.numeric(C_mat),
#   nrow = length(C_mat),
#   ncol = ncol(bulk_mat_t),
#   byrow = FALSE
# )

dim(ref_mat)
length(C_mat)
dim(bulk_mat_t)
length(w_vec)

undebug(deconvolute)
res <- deconvolute(ref_mat, bulk_mat_t, length_vec = C_mat, weight_vec = w_vec, lr = 0.01, epochs = 3000)


ComplexHeatmap::Heatmap(res$exposures, cluster_rows = F, cluster_columns = F,col = viridis_light_high)
```

```{r}
#debug(flscuts::iterative_LSI)


cor(cds@int_metadata$LSI_model$svd$d, seuHR@reductions$lsi@misc$svd$d)       # ≈ 1.00

identical(
  cds@int_metadata$LSI_model$features,
  seuHR@reductions$lsi@misc$features
)

identical(
  cds@int_metadata$LSI_model$features,
  VariableFeatures(seuHR)
)

library(SeuratDisk)
SeuratDisk::SaveH5Seurat(seuHR, filename = file.path("~/Desktop/seuHR.h5Seurat"))
SeuratDisk::Convert("~/Desktop/seuHR.h5Seurat", dest = "h5ad", overwrite = T)


```




```{r}
filter_calls <- function(seu_obj, training_result, prob_thresh = 0.95, pop_thresh = 0.05, labels, output_name = "tumor_call"){
  max <- apply(training_result$training_output$probs, 1, max)
  names(max) <- labels[apply(training_result$training_output$probs, 1, which.max)]
  names(max)[max<prob_thresh]<- NA
  seu_obj[[output_name]] <- names(max)
  tk <- names(table(names(max)))[table(names(max)) > pop_thresh * sum(table(names(max)))]
  seu_obj[[paste0(output_name, "_filtered")]]<- NA
  seu_obj[[paste0(output_name, "_filtered")]][names(max) %in% tk,1] <- names(max)[names(max) %in% tk]
  return(seu_obj)
}


```



```{r}
plot_gene_dispersion(seuHR)
seuHR <- select_features(seuHR, top_n = 5000, logmean_ul = -1, logmean_ll = -9)

vgHR <- get_selected_features(seuHR)

seuF <- calculate_feature_dispersion(seuF)
plot_gene_dispersion(seuF)
seuF <- select_features(seuF, top_n = 5000, logmean_ul = -3, logmean_ll = -10)
vgF <- get_selected_features(seuF)

seuF <- NormalizeData(seuF) %>% ScaleData()


res <- viewmastR(query_cds = seuHR, ref_cds = seuF, ref_celldata_col = "cell.labels", selected_features = vgF, return_probs = T, return_type = "list")
seuHR <- filter_calls(seuHR, prob_thresh = .9, pop_thresh = 0.001, training_result = res, labels = levels(factor(seuF$cell.labels)), output_name = "fetal")

DimPlot(seuHR, group.by = "fetal_filtered", cols = as.character(c(pals::alphabet(), pals::polychrome(), pals::glasbey())))


seu3 <- readRDS('/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchCancerCenter/Furlan_Lab - General/experiments/Peds_UCSF/cds/KMT2Apatients_seuratobject_cleaned.rds')

res3 <- viewmastR(query_cds = seu3, ref_cds = seuF, ref_celldata_col = "cell.labels", selected_features = vgF, return_probs = T, return_type = "list")
seu3 <- filter_calls(seu3, prob_thresh = .2, pop_thresh = 0.001, training_result = res3, labels = levels(factor(seuF$cell.labels)), output_name = "fetal")
DimPlot(seu3, group.by = "fetal_filtered", cols = as.character(c(pals::alphabet(), pals::polychrome(), pals::glasbey())))


plotly::ggplotly(DimPlot(seuHR, group.by = "viewmastR_pred", cols = as.character(c(pals::alphabet(), pals::polychrome(), pals::glasbey()))))

seuF <- viewmastR(seuF, seuHR, ref_celldata_col = "SFClassification", selected_features = vgHR)
seuF@meta.data
DimPlot(seuF, group.by = "viewmastR_pred", cols = cols)
FeaturePlot_scCustom(seuF, features = "FOXP3")
plotly::ggplotly(DimPlot(seuF, group.by = "cell.labels", raster = F, cols = as.character(c(pals::alphabet(), pals::polychrome(), pals::glasbey()))))
```


```{r}
colnames(seuHR@assays$RNA@data)

n_top <- c(3000,2500,2500) #number of features
seuHR <- flscuts::iterative_LSI(seuHR, num_dim = 25, num_features = n_top, resolution = c(2e-4, 9e-4, 1e-3), verbose = T, do_tf_idf = F)


undebug(flscuts::iterative_LSI)


single_cell_ref_column = "cell.labels"
bulk_count_data = file.path(ROOT_DIR3, "aml_all_bulk_counts.csv")
# bulk_meta_data = "~/Fred Hutchinson Cancer Research Center/Furlan_Lab - General/datasets/AML/1031/coembed/data/bulk_meta.csv" #don't really need this
gene_weights = file.path(ROOT_DIR2, "geneWeights.tsv")


seuHR <- file.path()
```

```{r}
library(Seurat)
seu <- readRDS(single_cell_ref)
plotly::ggplotly(DimPlot(seu, group.by = "cell.labels", raster = F, cols = as.character(c(pals::alphabet(), pals::polychrome(), pals::glasbey()))))

```



## Running viewmastR

```{r, dpi=300, fig.height=4, fig.width = 6}
suppressPackageStartupMessages({
library(viewmastR)
library(Seurat)
library(ggplot2)
})

#query dataset
seu <- readRDS(file.path(ROOT_DIR1, "240813_final_object.RDS"))
#reference dataset
seur<-readRDS(file.path(ROOT_DIR2, "230329_rnaAugmented_seurat.RDS"))
vg <- get_selected_features(seu)
```

## Build the model and infer for small dataset (not using chunks and parallelization)

This is also covered elsewhere
```{r}
results <- viewmastR(seu, seur, ref_celldata_col = "SFClassification", selected_features = vg, max_epochs = 4, train_only = T)



seu<-viewmastR_infer(seu, results[["model_dir"]], vg, labels = levels(factor(seur$SFClassification)), batch_size = 250)
DimPlot(seu, group.by = "viewmastR_inferred", cols = seur@misc$colors)
```

## Build the model and infer for large dataset (dividing the query into chunks and using parallelization)

By using chunks and workers, you can infer from the model only chunks at a time using multiple workers in parallel.
```{r}
options(future.globals.maxSize = 16 * 1024^3)
seu<-viewmastR_infer(seu, results[["model_dir"]], query_celldata_col = "viewmastR_inferred_parallel", vg, labels = levels(factor(seur$SFClassification)), chunks = 4, workers = 4)
DimPlot(seu, group.by = "viewmastR_inferred_parallel", cols = seur@misc$colors)
```

We see no difference if parallelization is used
```{r, dpi=300, fig.height=6, fig.width=10}
confusion_matrix(pred = factor(seu$viewmastR_inferred), gt = factor(seu$viewmastR_inferred_parallel), cols = seur@misc$colors)
```


## Bigger dataset
By optimising chunk size, workers and batch size, you can achieve faster results on big datasets. Here we are just inferring back on the larger reference set (because it is larger, but this is not advisable in normal practice)
```{r}
seur<-viewmastR_infer(seur, results[["model_dir"]], query_celldata_col = "viewmastR_inferred_parallel", vg, labels = levels(factor(seur$SFClassification)), chunks = 20, workers = 16, batch_size = 3000)
DimPlot(seur, group.by = "viewmastR_inferred_parallel", cols = seur@misc$colors)
```

```{r, dpi=300, fig.height=6, fig.width=10}
confusion_matrix(pred = factor(seur$SFClassification), gt = factor(seur$viewmastR_inferred_parallel), cols = seur@misc$colors)
```

