---
title: "How to use viewmastR to deconvolute"
output: html_document
date: "2024-12-10"
always_allow_html: true
editor_options: 
  chunk_output_type: console
---


```{r, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
rm(list=ls())


if(grepl("^gizmo", Sys.info()["nodename"])){
  ROOT_DIR1<-"/fh/fast/furlan_s/experiments/MB_10X_5p/cds"
  ROOT_DIR2<-"/fh/fast/furlan_s/grp/data/ddata/BM_data"
} else {
  ROOT_DIR1<-"/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchinsonCancerCenter/Furlan_Lab - General/experiments/MB_10X_5p/cds"
  ROOT_DIR2<-"/Users/sfurlan/Fred Hutch Cancer Center/Furlan_Lab - General/datasets/fetalBM/data"
  ROOT_DIR3 <- "/Users/sfurlan/Fred Hutch Cancer Center/Furlan_Lab - General/datasets/AML/1031/coembed/data"
  ROOT_DIR4 <- "/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchinsonCancerCenter/Furlan_Lab - General/datasets/Healthy_BM_greenleaf"
}
seurat_object_loc <- "/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchCancerCenter/Furlan_Lab - General/datasets/Healthy_BM_greenleaf/230329_rnaAugmented_seurat.RDS"
```

#### Load a seurat object

This is an atlas of healthy bone marrow celltypes taken from:

Granja JM, Klemm S, McGinnis LM, Kathiria AS, Mezger A, Corces MR, Parks B, Gars E, Liedtke M, Zheng GXY, Chang HY, Majeti R, Greenleaf WJ. Single-cell multiomic analysis identifies regulatory programs in mixed-phenotype acute leukemia. Nat Biotechnol. 2019 Dec;37(12):1458-1465. doi: 10.1038/s41587-019-0332-7. Epub 2019 Dec 2. PMID: 31792411; PMCID: PMC7258684.

```{r}
library(Seurat)
library(viewmastR)
# Load single-cell data
sc_data <- readRDS(seurat_object_loc)
DimPlot(sc_data, group.by = "SFClassification", cols = sc_data@misc$colors)
```

Now we pseudobulk the samples to get their "signatures"

```{r}
# Extract normalized counts and cell type labels
counts <- GetAssayData(sc_data, layer = "data")  # log-normalized
cell_types <- sc_data$SFClassification


# Aggregate by cell type and create signatures
signatures <- sapply(unique(cell_types), function(ct) {
  cells <- which(cell_types == ct)
  rowMeans(expm1(counts[, cells]))  # Convert back from log
})

# Normalize to proportions (sum to 1)
signatures <- sweep(signatures, 2, colSums(signatures), "/")

# or load the signatures from a file
signatures <- "/Users/sfurlan/Desktop/scSigs.csv"
signatures <- read.csv(signatures)
rn <- signatures$var_names
signatures <- as.matrix(signatures[,2:ncol(signatures)])
rownames(signatures)<-rn
colnames(signatures) <- substr(colnames(signatures), 2, nchar(colnames(signatures)))

# Load bulk counts and gene lengths
bulk_counts <- "/Users/sfurlan/Library/CloudStorage/OneDrive-FredHutchCancerCenter/computation/Analysis/ddata/MPAL/data/bulk/GSE74246_RNAseq_All_Counts.txt"
bulk_counts <- read.csv(bulk_counts, sep = "\t")
rn <- bulk_counts$X_TranscriptID
bulk_counts$X_TranscriptID<- NULL
bulk_counts <- as.matrix(bulk_counts)
rownames(bulk_counts) <- rn
gene_info <- read.table("/Users/sfurlan/Fred Hutch Cancer Center/Furlan_Lab - General/datasets/fetalBM/data/geneWeights_withSymbol.tsv", header = TRUE)
rownames(gene_info) <- gene_info$symbol

# Ensure same genes
common_genes <- intersect(intersect(rownames(signatures), rownames(bulk_counts)), gene_info$symbol)
signatures <- signatures[common_genes, ]
bulk_counts <- bulk_counts[common_genes, ]

# Get gene lengths
gene_lengths <- gene_info[common_genes, "length"]
gene_weights <- gene_info[common_genes, "weight"]

dim(signatures)
dim(bulk_counts)
length(gene_lengths)

# #fast
#   # user  system elapsed 
#   # 3.852   0.809   9.530 
# system.time(result <- deconvolve_bulk(
#   signatures = signatures,
#   bulk_counts = bulk_counts+0.00001,
#   gene_lengths = gene_lengths+0.00001,
#   init_log_exp = -10,  # Start with higher initial values
#   learning_rate = 0.1,  # Increase learning rate
#   max_iter = 2000,
#   backend = "wgpu"
# ))
# 
# #long 
#  #   user  system elapsed 
#  # 36.121   6.782  39.196 
# 
# system.time(result <- deconvolve_bulk(
#   signatures = signatures,
#   bulk_counts = bulk_counts+0.00001,
#   gene_lengths = gene_lengths+0.00001,
#   init_log_exp = -10,  # Start with higher initial values
#   learning_rate = 0.1,  # Increase learning rate
#   max_iter = 2000,
#   backend = "ndarray"
# ))
# 
# #very very long
# #    user  system elapsed 
# # 199.663 162.284 319.761 
# system.time(result <- deconvolve_bulk(
#   signatures = signatures,
#   bulk_counts = bulk_counts+0.00001,
#   gene_lengths = gene_lengths+0.00001,
#   init_log_exp = -10,  # Start with higher initial values
#   learning_rate = 0.1,  # Increase learning rate
#   max_iter = 2000,
#   backend = "candle"
# ))

```

## tuning
```{r}
undebug(deconvolve_bulk)
debug(calculate_fit_metrics)
system.time(
result <- deconvolve_bulk(
  signatures = signatures,
  bulk_counts = bulk_counts,
  gene_lengths = gene_lengths,
  gene_weights = gene_weights,
  poll_interval = 300,
  insert_size = 500,
  init_log_exp = -3,     # CHANGED: Start higher
  learning_rate = 0.01,
  l1_lambda = 100,       # CHANGED: Start smaller  
  l2_lambda = 0,
  max_iter = 5000,       
  backend = "wgpu"
)
)

 #   user  system elapsed 
 # 51.647   9.889 119.599 


cor(as.vector(result$pred_counts[gene_weights>0,]), as.vector(bulk_counts[gene_weights>0,]))

obs_flat <- as.vector(bulk_counts[gene_weights>0,])
pred_flat <- as.vector(result$pred_counts[gene_weights>0,])

# Remove zeros for log scale
keep_idx <- obs_flat > 0 & pred_flat > 0

fit_df <- data.frame(
  Observed = log1p(obs_flat[keep_idx]),
  Predicted = log1p(pred_flat[keep_idx])
)
    
    
ggplot(fit_df, aes(x=Observed, y=Predicted))+geom_hex(bins = 100)+scale_fill_viridis_c(trans = "log10")

print_metrics_summary(result)

plot_deconvolution(result, observed_counts = bulk_counts, type = "all")

```


```{r}
d <- result0$proportions_with_intercept
d <- d[,!(grepl("LSC", colnames(d)) | grepl("rHSC", colnames(d)) | grepl("Blast", colnames(d)))]
mr <- as.matrix(d)
sm <- scale(mr)
cols <- pals::glasbey(13)
names(cols) <- unique(strsplit(colnames(mr), "\\.") %>% sapply("[[", 2))
cola <- HeatmapAnnotation(bulk = strsplit(colnames(mr), "\\.") %>% sapply("[[", 2), col = list(bulk=cols))
Heatmap(mr, top_annotation = cola, show_column_names = F, col = viridis_light_high)

```



```{r}
library(ggplot2)
m_python <- m_python[,colnames(mr)]

#seuHR <- readRDS("/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchCancerCenter/Furlan_Lab - General/datasets/Healthy_BM_greenleaf/230329_rnaAugmented_seurat.RDS")
ggplot(data.frame(Rust=as.vector(mr), Python=as.vector(m_python), Decon = rep(rownames(mr), ncol(mr))), aes(x=Python, y=Rust, colour = Decon))+geom_point()+scale_color_manual(values = seuHR@misc$colors)


cor(data.frame(Rust=as.vector(mr), Python=as.vector(m_python)))

```


```{r}
# Find the worst-predicted genes
pred_full_matrix <- result$pred_counts
obs_matrix <- bulk_counts

# Calculate residuals for each gene across all samples
residuals <- obs_matrix - pred_full_matrix
abs_residuals <- abs(residuals)
mean_abs_resid <- rowMeans(abs_residuals)

# Find worst genes
worst_genes_idx <- order(mean_abs_resid, decreasing = TRUE)[1:20]

cat("Top 20 worst-predicted genes:\n")
for (i in worst_genes_idx[1:10]) {
  cat(sprintf("Gene %d: mean_obs=%.0f, mean_pred=%.0f, mean_|error|=%.0f\n",
              i, 
              mean(obs_matrix[i,]), 
              mean(pred_full_matrix[i,]),
              mean_abs_resid[i]))
}

# Check if these are weighted zero
cat("\nGene weights for worst genes:\n")
print(gene_weights[worst_genes_idx[1:10]])

# Plot residuals vs observed for worst gene
worst_gene <- worst_genes_idx[1]
plot(obs_matrix[worst_gene,], residuals[worst_gene,],
     main = paste("Residuals for worst gene", worst_gene),
     xlab = "Observed", ylab = "Residual (obs - pred)")
abline(h = 0, col = "red")
```

```{r}
# 1. Are signature column sums exactly 1?
colSums(signatures)

# 2. Which cell types have the lowest mean expression in signatures?
colMeans(signatures)

# 3. Are there any cell types with extreme values?
apply(signatures, 2, function(x) c(min=min(x), max=max(x), nonzero=mean(x>0)))

# 4. What do the raw exposures look like for the problematic cell types?
# Compare pDC (cell type 9) and B cells (cell type 14)
rust_pdc <- result$exposures[9,]
rust_b <- result$exposures[14,]
summary(rust_pdc)
summary(rust_b)

```

```{r}
# Check raw exposures
summary(result$exposures)

# Check if intercept dominates
mean(result$intercept_fraction)

# Check predicted vs observed counts
cor(as.vector(result$pred_counts), as.vector(bulk_counts))
plot(log1p(as.vector(bulk_counts)), log1p(as.vector(result$pred_counts)))


```

## testing scaled regularization
```{r}
result4 <- deconvolve_bulk(
  signatures = signatures,
  bulk_counts = bulk_counts,
  gene_lengths = gene_lengths,
  gene_weights = gene_weights,
  poll_interval = 300,
  init_log_exp = -10,     # CHANGED: Start higher
  learning_rate = 0.01,
  l1_lambda = 1e-4,       # CHANGED: Start smaller  
  l2_lambda = 0,
  max_iter = 5000,       
  backend = "wgpu"
)

result3 <- deconvolve_bulk(
  signatures = signatures,
  bulk_counts = bulk_counts,
  gene_lengths = gene_lengths,
  gene_weights = gene_weights,
  poll_interval = 300,
  init_log_exp = -10,     # CHANGED: Start higher
  learning_rate = 0.01,
  l1_lambda = 1e-3,       # CHANGED: Start smaller  
  l2_lambda = 0,
  max_iter = 5000,       
  backend = "wgpu"
)
result2 <- deconvolve_bulk(
  signatures = signatures,
  bulk_counts = bulk_counts,
  gene_lengths = gene_lengths,
  gene_weights = gene_weights,
  poll_interval = 300,
  init_log_exp = -10,     # CHANGED: Start higher
  learning_rate = 0.01,
  l1_lambda = 1e-2,       # CHANGED: Start smaller  
  l2_lambda = 0,
  max_iter = 5000,       
  backend = "wgpu"
)
result1 <- deconvolve_bulk(
  signatures = signatures,
  bulk_counts = bulk_counts,
  gene_lengths = gene_lengths,
  gene_weights = gene_weights,
  poll_interval = 300,
  init_log_exp = -10,     # CHANGED: Start higher
  learning_rate = 0.01,
  l1_lambda = 1e-1,       # CHANGED: Start smaller  
  l2_lambda = 0,
  max_iter = 5000,       
  backend = "wgpu"
)
result0 <- deconvolve_bulk(
  signatures = signatures,
  bulk_counts = bulk_counts,
  gene_lengths = gene_lengths,
  gene_weights = gene_weights,
  poll_interval = 300,
  init_log_exp = -10,     # CHANGED: Start higher
  learning_rate = 0.01,
  l1_lambda = 5,       # CHANGED: Start smaller  
  l2_lambda = 0,
  max_iter = 5000,       
  backend = "wgpu"
)


result01 <- deconvolve_bulk(
  signatures = signatures,
  bulk_counts = bulk_counts,
  gene_lengths = gene_lengths,
  gene_weights = gene_weights,
  poll_interval = 300,
  init_log_exp = -10,     # CHANGED: Start higher
  learning_rate = 0.01,
  l1_lambda = 100,       # CHANGED: Start smaller  
  l2_lambda = 0,
  max_iter = 5000,       
  backend = "wgpu"
)


# Compare
compare_results <- function(r1, r2, name1, name2) {
  n1 <- sum(rowMeans(r1$exposures) > 0.01)
  n2 <- sum(rowMeans(r2$exposures) > 0.01)
  cat(sprintf("%s: %d non-zero cell types, mean_exp=%.1f\n", 
              name1, n1, mean(r1$exposures)))
  cat(sprintf("%s: %d non-zero cell types, mean_exp=%.1f\n", 
              name2, n2, mean(r2$exposures)))
  cat(sprintf("Difference: %d cell types\n\n", n1 - n2))
}

compare_results(result4, result4, "L1=0.0001", "L1=0.001")
compare_results(result3, result2, "L1=0.001", "L1=0.01")
compare_results(result2, result1, "L1=0.01", "L1=0.1")
compare_results(result1, result0, "L1=0.1", "L1=1")
compare_results(result0, result01, "L1=1", "L1=100")
```

#Redo this in R

```{r}
library(viewmastR)


seuHR <- readRDS("/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchCancerCenter/Furlan_Lab - General/datasets/Healthy_BM_greenleaf/230329_rnaAugmented_seurat.RDS")
DimPlot(seuHR, group.by = "SFClassification")
bulk_count_data = "/Users/sfurlan/Library/CloudStorage/OneDrive-FredHutchCancerCenter/computation/Analysis/ddata/MPAL/data/bulk/GSE74246_RNAseq_All_Counts.txt"
gene_weights = "/Users/sfurlan/Fred Hutch Cancer Center/Furlan_Lab - General/datasets/fetalBM/data/geneWeights_withSymbol.tsv"
fit_csv = "/Users/sfurlan/Desktop/test_fit_rust.csv"


bulk_count_data <- read.csv(bulk_count_data, sep = "\t")
rn <- bulk_count_data$X_TranscriptID
bulk_count_data$X_TranscriptID<- NULL
bulk_count_data <- as.matrix(bulk_count_data)
rownames(bulk_count_data) <- rn
gene_weights <- read.table(gene_weights, sep = "\t", header = T)

pseudobulk <- function(seu, by, assay = "RNA", layer = "data"){
  meta <- seu@meta.data
  counts_mat <- GetAssayData(seu, assay = assay, layer = layer)
  # extract your cell‐type factor
  cell_types <- as.character(meta[[by]])
  unique_types <- unique(cell_types)
  
  # pre‐allocate an empty matrix: genes × n_types
  genes       <- rownames(counts_mat)
  scSigs_mat  <- matrix(0,
                        nrow = length(genes),
                        ncol = length(unique_types),
                        dimnames = list(genes, unique_types))
  
  # fill it
  for (ct in unique_types) {
    idx <- which(cell_types == ct)
    # sum counts for that cell-type
    vec <- rowSums(counts_mat[, idx, drop=FALSE])
    # normalize so they sum to 1
    scSigs_mat[, ct] <- vec / sum(vec)
  }
  scSigs_mat
}


undebug(pseudobulk)
# 1) build healthy reference from your non-malignant clusters
ref_mat <- pseudobulk(seuHR, "SFClassification" )
bulk_mat <- bulk_count_data
C_mat <- gene_weights$length
# C_mat <- bulk_count_data$size
names(C_mat)<-gene_weights$symbol
# bulk_mat <- as.matrix(bulk_count_data[,3:ncol(bulk_count_data)])
# rownames(bulk_mat)<-bulk_count_data$X

common_feats <- intersect(intersect(rownames(ref_mat), names(C_mat)), rownames(bulk_mat))


ref_mat <- ref_mat[common_feats, ]
C_mat <- C_mat[common_feats]
bulk_mat <- bulk_mat[common_feats,]
w_vec <- gene_weights$weight[match(common_feats, gene_weights$symbol)]

bulk_mat_t <- bulk_mat
dim(bulk_mat_t)

saveRDS(list(ref_mat, bulk_mat_t, C_mat, w_vec), file.path("~/Desktop/seuHR/test_list.RDS"))



library(viewmastR)

l<-readRDS("~/Desktop/seuHR/test_list.RDS")
ref_mat = l[[1]]
bulk_mat_t = l[[2]]
C_mat = l[[3]]
w_vec <- l[[4]]


result <- res$exposures
dim(result)
rownames(result) <- c(colnames(ref_mat), "Intercept")
colnames(result) <- colnames(bulk_mat_t)
d <- result
d <- d[,!(grepl("LSC", colnames(d)) | grepl("rHSC", colnames(d)) | grepl("Blast", colnames(d)))]
m <- as.matrix(d)
cols <- pals::glasbey(13)
names(cols) <- unique(strsplit(colnames(m), "\\.") %>% sapply("[[", 2))
cola <- HeatmapAnnotation(bulk = strsplit(colnames(m), "\\.") %>% sapply("[[", 2), col = list(bulk=cols))
Heatmap(m, bottom_annotation = cola, show_column_names = F, col = viridis_light_high)




```

```{r}

cds <- readRDS(file.path(ROOT_DIR4, "healthy_bm_greenleaf.rds"))
plot_cells(cds, color_cells_by = "BioClassification")

rownames(cds)<-rowData(cds)$gene_id
#cds <- normalized_counts(cds)
PCs <- 25 #Number of PCs for clustering
n_top <- c(3000,2500,2500) #number of features
cds<-iterative_LSI(cds, num_dim = PCs, num_features = n_top, resolution = c(2e-4, 9e-4, 1e-3), verbose = T)
cds<-reduce_dimension(cds, reduction_method = "UMAP", preprocess_method = "LSI", umap.min_dist = 0.45,  umap.n_neighbors = 35, umap.metric = "euclidean", verbose=T)
plot_cells(cds, color_cells_by = "BioClassification")


feats <- cds@int_metadata$LSI_model$features

#make a seurat object for export
counts <- cds@assays@data$counts
rownames(counts)<-rowData(cds)$gene_id
seuHR <- CreateSeuratObject(counts = counts , meta.data = data.frame(colData(cds)))
seuHR <- NormalizeData(seuHR) %>% ScaleData()
lsimat <- cds@int_metadata$LSI_model$svd$v
rownames(lsimat)<-colnames(cds)
colnames(lsimat)<-paste0("LSI_", 1:ncol(lsimat))
seuHR$lsi <- CreateDimReducObject(embeddings = lsimat, assay = "RNA")
# seuHR <- readRDS(file.path(ROOT_DIR4, "healthy_bm_greenleaf.rds"))
# seuHR <- CreateSeuratObject(counts = seuHR@assays@data$counts, meta.data = data.frame(colData(seuHR)))
# seuHR <- NormalizeData(seuHR) %>% ScaleData()
# VariableFeatures(seuHR)<- feats
# seuHR <- Signac::RunTFIDF(seuHR)
# seuHR <- Signac::RunSVD(seuHR)
VariableFeatures(seuHR)<-feats
seuHR <- RunUMAP(seuHR, dims = 1:25, reduction = "lsi", min.dist = 0.45, n.neighbors = 35, metric = "euclidean")
DimPlot(seuHR, group.by = "BioClassification")
DimPlot(seuHR, group.by = "Group")

dir.create("~/Desktop/seuHR")
flscuts::export_seurat_to_10x(seuHR, dir = "~/Desktop/seuHR")
```




```{r}

library(viewmastR)

bulk_count_data = "/Users/sfurlan/Fred Hutch Cancer Center/Furlan_Lab - General/datasets/AML/1031/coembed/data/aml_all_bulk_counts.csv"
gene_weights = "/Users/sfurlan/Fred Hutch Cancer Center/Furlan_Lab - General/datasets/fetalBM/data/geneWeights.tsv"

bulk_count_data <- read.csv(bulk_count_data)
gene_weights <- read.table(gene_weights, sep = "\t", header = T)

pseudobulk <- function(seu, by, assay = "RNA", layer = "counts"){
  meta <- seu@meta.data
  counts_mat <- GetAssayData(seu, assay = assay, layer = layer)
  # extract your cell‐type factor
  cell_types <- as.character(meta[[by]])
  unique_types <- unique(cell_types)
  
  # pre‐allocate an empty matrix: genes × n_types
  genes       <- rownames(counts_mat)
  scSigs_mat  <- matrix(0,
                        nrow = length(genes),
                        ncol = length(unique_types),
                        dimnames = list(genes, unique_types))
  
  # fill it
  for (ct in unique_types) {
    idx <- which(cell_types == ct)
    # sum counts for that cell-type
    vec <- rowSums(counts_mat[, idx, drop=FALSE])
    # normalize so they sum to 1
    scSigs_mat[, ct] <- vec / sum(vec)
  }
  scSigs_mat
}



# 1) build healthy reference from your non-malignant clusters
ref_mat <- pseudobulk(seuHR, "BioClassification" )
C_mat <- bulk_count_data$size
names(C_mat)<-bulk_count_data$X
bulk_mat <- as.matrix(bulk_count_data[,3:ncol(bulk_count_data)])
rownames(bulk_mat)<-bulk_count_data$X

common_feats <- intersect(intersect(rownames(ref_mat), names(C_mat)), gene_weights$geneID)


ref_mat <- ref_mat[common_feats, ]
C_mat <- C_mat[common_feats]
bulk_mat <- bulk_mat[common_feats,]
w_vec <- gene_weights$weight[match(common_feats, gene_weights$geneID)]


tester <- c(2116, 3432, 3391, 963, 4374, 1277, 3943, 1295, 3714, 3576, 1263, 2275, 583, 2497, 889, 1380, 4483, 4431, 3146, 1955, 1997, 133, 3573, 3622, 660, 2902, 30, 1017, 2783, 1919, 2996, 4908, 2051, 1474, 1018, 631, 180, 3758, 2229, 986, 2302, 2197, 3451, 3267, 2347, 4867, 2328, 2638, 2827, 737, 371, 4592, 647, 3349, 3562, 3482, 2389, 489, 3046, 479, 2293, 1079, 2131, 4072, 1907, 3456, 2255, 540, 4294, 3011, 4610, 2970, 4819, 3088, 3114, 584, 192, 2751, 945, 1158, 1696, 3889, 1020, 2533, 1579, 4598, 3182, 2362, 1580, 2713, 3603, 4635, 3678, 2451, 2332, 4, 2168, 3370, 3502, 305)

bulk_mat_t <- bulk_mat[,tester]
dim(bulk_mat_t)

#saveRDS(list(ref_mat, bulk_mat_t, C_mat, w_vec), file.path("~/Desktop/seuHR/test_list.RDS"))



library(viewmastR)

l<-readRDS("~/Desktop/seuHR/test_list.RDS")
ref_mat = l[[1]]
bulk_mat_t = l[[2]]
C_mat = l[[3]]
w_vec <- l[[4]]

deconvolute <- function(
    ref_mat, 
    bulk_mat, 
    length_vec, 
    weight_vec,
    init_log_exp = -5,
    lr           = 1e-3,
    l1_lambda    = 0.01,
    l2_lambda    = 0.01,
    epochs       = 300)
{
    fit <- fit_deconv(
      sigs = ref_mat,
      bulk = bulk_mat,
      mol2read = length_vec,
      w_vec = weight_vec,
      init_log_exp =init_log_exp,
      lr          = lr,
      l1_lambda   = l1_lambda,
      l2_lambda   = l2_lambda,
      epochs      = epochs)
  
        ## -- 1. sort the sample columns the way pandas .sort_index(axis=1) does ----
      loss <- fit$loss
      fit <- fit$exposures[ , order(colnames(bulk_mat))]
      # 
      # 
      # ## -- 2. split out “goodness-of-fit” rows -----------------------------------
      # gof_noms <- c("pR2", "fitCount", "obsCount")
      # gof <- fit[gof_noms, , drop = FALSE]
      # 
      # ## add log2(countRatio) = log2(fitCount / obsCount) per sample
      # gof["log2(countRatio)", ] <- log2(gof["fitCount", ] / gof["obsCount", ])
      # 
      # ## -- 3. raw exposures (everything that is *not* a GOF row) -----------------
      # expos_idx <- !rownames(fit) %in% gof_noms
      # raw <- fit[expos_idx, , drop = FALSE]
    
      ## -- 4. column-wise normalisation so each sample sums to 1 -----------------
      exposures <- sweep(fit, 2, colSums(fit), FUN = "/")
    
      ## -- 5. return a list mimicking the Python dict ----------------------------
      list(
        exposures = exposures
        # gof       = gof,
        # raw       = raw          # raw (unnormalised) exposure matrix
      )
}

# C_mat <- matrix(
#   as.numeric(C_mat),
#   nrow = length(C_mat),
#   ncol = ncol(bulk_mat_t),
#   byrow = FALSE
# )

dim(ref_mat)
length(C_mat)
dim(bulk_mat_t)
length(w_vec)

undebug(deconvolute)
res <- deconvolute(ref_mat, bulk_mat_t, length_vec = C_mat, weight_vec = w_vec, lr = 0.01, epochs = 3000)


ComplexHeatmap::Heatmap(res$exposures, cluster_rows = F, cluster_columns = F,col = viridis_light_high)
```

```{r}
#debug(flscuts::iterative_LSI)


cor(cds@int_metadata$LSI_model$svd$d, seuHR@reductions$lsi@misc$svd$d)       # ≈ 1.00

identical(
  cds@int_metadata$LSI_model$features,
  seuHR@reductions$lsi@misc$features
)

identical(
  cds@int_metadata$LSI_model$features,
  VariableFeatures(seuHR)
)

library(SeuratDisk)
SeuratDisk::SaveH5Seurat(seuHR, filename = file.path("~/Desktop/seuHR.h5Seurat"))
SeuratDisk::Convert("~/Desktop/seuHR.h5Seurat", dest = "h5ad", overwrite = T)


```




```{r}
filter_calls <- function(seu_obj, training_result, prob_thresh = 0.95, pop_thresh = 0.05, labels, output_name = "tumor_call"){
  max <- apply(training_result$training_output$probs, 1, max)
  names(max) <- labels[apply(training_result$training_output$probs, 1, which.max)]
  names(max)[max<prob_thresh]<- NA
  seu_obj[[output_name]] <- names(max)
  tk <- names(table(names(max)))[table(names(max)) > pop_thresh * sum(table(names(max)))]
  seu_obj[[paste0(output_name, "_filtered")]]<- NA
  seu_obj[[paste0(output_name, "_filtered")]][names(max) %in% tk,1] <- names(max)[names(max) %in% tk]
  return(seu_obj)
}


```



```{r}
plot_gene_dispersion(seuHR)
seuHR <- select_features(seuHR, top_n = 5000, logmean_ul = -1, logmean_ll = -9)

vgHR <- get_selected_features(seuHR)

seuF <- calculate_feature_dispersion(seuF)
plot_gene_dispersion(seuF)
seuF <- select_features(seuF, top_n = 5000, logmean_ul = -3, logmean_ll = -10)
vgF <- get_selected_features(seuF)

seuF <- NormalizeData(seuF) %>% ScaleData()


res <- viewmastR(query_cds = seuHR, ref_cds = seuF, ref_celldata_col = "cell.labels", selected_features = vgF, return_probs = T, return_type = "list")
seuHR <- filter_calls(seuHR, prob_thresh = .9, pop_thresh = 0.001, training_result = res, labels = levels(factor(seuF$cell.labels)), output_name = "fetal")

DimPlot(seuHR, group.by = "fetal_filtered", cols = as.character(c(pals::alphabet(), pals::polychrome(), pals::glasbey())))


seu3 <- readRDS('/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchCancerCenter/Furlan_Lab - General/experiments/Peds_UCSF/cds/KMT2Apatients_seuratobject_cleaned.rds')

res3 <- viewmastR(query_cds = seu3, ref_cds = seuF, ref_celldata_col = "cell.labels", selected_features = vgF, return_probs = T, return_type = "list")
seu3 <- filter_calls(seu3, prob_thresh = .2, pop_thresh = 0.001, training_result = res3, labels = levels(factor(seuF$cell.labels)), output_name = "fetal")
DimPlot(seu3, group.by = "fetal_filtered", cols = as.character(c(pals::alphabet(), pals::polychrome(), pals::glasbey())))


plotly::ggplotly(DimPlot(seuHR, group.by = "viewmastR_pred", cols = as.character(c(pals::alphabet(), pals::polychrome(), pals::glasbey()))))

seuF <- viewmastR(seuF, seuHR, ref_celldata_col = "SFClassification", selected_features = vgHR)
seuF@meta.data
DimPlot(seuF, group.by = "viewmastR_pred", cols = cols)
FeaturePlot_scCustom(seuF, features = "FOXP3")
plotly::ggplotly(DimPlot(seuF, group.by = "cell.labels", raster = F, cols = as.character(c(pals::alphabet(), pals::polychrome(), pals::glasbey()))))
```


```{r}
colnames(seuHR@assays$RNA@data)

n_top <- c(3000,2500,2500) #number of features
seuHR <- flscuts::iterative_LSI(seuHR, num_dim = 25, num_features = n_top, resolution = c(2e-4, 9e-4, 1e-3), verbose = T, do_tf_idf = F)


undebug(flscuts::iterative_LSI)


single_cell_ref_column = "cell.labels"
bulk_count_data = file.path(ROOT_DIR3, "aml_all_bulk_counts.csv")
# bulk_meta_data = "~/Fred Hutchinson Cancer Research Center/Furlan_Lab - General/datasets/AML/1031/coembed/data/bulk_meta.csv" #don't really need this
gene_weights = file.path(ROOT_DIR2, "geneWeights.tsv")


seuHR <- file.path()
```

```{r}
library(Seurat)
seu <- readRDS(single_cell_ref)
plotly::ggplotly(DimPlot(seu, group.by = "cell.labels", raster = F, cols = as.character(c(pals::alphabet(), pals::polychrome(), pals::glasbey()))))

```



## Running viewmastR

```{r, dpi=300, fig.height=4, fig.width = 6}
suppressPackageStartupMessages({
library(viewmastR)
library(Seurat)
library(ggplot2)
})

#query dataset
seu <- readRDS(file.path(ROOT_DIR1, "240813_final_object.RDS"))
#reference dataset
seur<-readRDS(file.path(ROOT_DIR2, "230329_rnaAugmented_seurat.RDS"))
vg <- get_selected_features(seu)
```

## Build the model and infer for small dataset (not using chunks and parallelization)

This is also covered elsewhere
```{r}
results <- viewmastR(seu, seur, ref_celldata_col = "SFClassification", selected_features = vg, max_epochs = 4, train_only = T)



seu<-viewmastR_infer(seu, results[["model_dir"]], vg, labels = levels(factor(seur$SFClassification)), batch_size = 250)
DimPlot(seu, group.by = "viewmastR_inferred", cols = seur@misc$colors)
```

## Build the model and infer for large dataset (dividing the query into chunks and using parallelization)

By using chunks and workers, you can infer from the model only chunks at a time using multiple workers in parallel.
```{r}
options(future.globals.maxSize = 16 * 1024^3)
seu<-viewmastR_infer(seu, results[["model_dir"]], query_celldata_col = "viewmastR_inferred_parallel", vg, labels = levels(factor(seur$SFClassification)), chunks = 4, workers = 4)
DimPlot(seu, group.by = "viewmastR_inferred_parallel", cols = seur@misc$colors)
```

We see no difference if parallelization is used
```{r, dpi=300, fig.height=6, fig.width=10}
confusion_matrix(pred = factor(seu$viewmastR_inferred), gt = factor(seu$viewmastR_inferred_parallel), cols = seur@misc$colors)
```


## Bigger dataset
By optimising chunk size, workers and batch size, you can achieve faster results on big datasets. Here we are just inferring back on the larger reference set (because it is larger, but this is not advisable in normal practice)
```{r}
seur<-viewmastR_infer(seur, results[["model_dir"]], query_celldata_col = "viewmastR_inferred_parallel", vg, labels = levels(factor(seur$SFClassification)), chunks = 20, workers = 16, batch_size = 3000)
DimPlot(seur, group.by = "viewmastR_inferred_parallel", cols = seur@misc$colors)
```

```{r, dpi=300, fig.height=6, fig.width=10}
confusion_matrix(pred = factor(seur$SFClassification), gt = factor(seur$viewmastR_inferred_parallel), cols = seur@misc$colors)
```

