---
title: "How to use viewmastR to deconvolute"
output: html_document
date: "2024-12-10"
always_allow_html: true
editor_options: 
  chunk_output_type: console
---


```{r, echo=FALSE, message=FALSE, warning=FALSE}
rm(list=ls())


if(grepl("^gizmo", Sys.info()["nodename"])){
  ROOT_DIR1<-"/fh/fast/furlan_s/experiments/MB_10X_5p/cds"
  ROOT_DIR2<-"/fh/fast/furlan_s/grp/data/ddata/BM_data"
} else {
  ROOT_DIR1<-"/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchinsonCancerCenter/Furlan_Lab - General/experiments/MB_10X_5p/cds"
  ROOT_DIR2<-"/Users/sfurlan/Fred Hutch Cancer Center/Furlan_Lab - General/datasets/fetalBM/data"
  ROOT_DIR3 <- "/Users/sfurlan/Fred Hutch Cancer Center/Furlan_Lab - General/datasets/AML/1031/coembed/data"
  ROOT_DIR4 <- "/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchinsonCancerCenter/Furlan_Lab - General/datasets/Healthy_BM_greenleaf"
}
library(Seurat)
library(scCustomize)
library(monocle3)
library(magrittr)
library(flscuts)


```


## Make a harmonized ENS (without version), Symbol, lengths and Weights dataset for deconvolution
```{r}
## start with the 10X gtf file
gff<-"/Users/sfurlan/refs/GRCh38/genes.gtf"
gff <- ape::read.gff(gff, na.strings = c(".", "?"), GFF3 = F)

genes <- gff[gff$feature=="gene",]
library(purrr)
library(stringr)


attr_pattern <- '(\\w+)\\s+"([^"]*)"|(\\w+)\\s+([^;"]+)'

parse_attributes_row <- function(s) {
  m <- stringr::str_match_all(s, attr_pattern)[[1]]
  if (nrow(m) == 0) return(tibble())  # empty row guard

  keys  <- ifelse(!is.na(m[,2]), m[,2], m[,4])
  vals  <- ifelse(!is.na(m[,3]), m[,3], m[,5])

  # clean
  keys <- trimws(keys)
  vals <- trimws(vals)

  # collapse duplicates (e.g., multiple 'tag' entries)
  collapsed <- lapply(split(vals, keys), function(x) paste(unique(x), collapse = ","))

  # return a one-row tibble with named fields
  tibble(!!!collapsed)
}

# Suppose your data frame is df with a column `attributes` like the example you pasted
attrs_wide <- genes %>%
  mutate(.parsed = map(attributes, parse_attributes_row)) %>%
  unnest_wider(.parsed, names_repair = "unique") %>%  # handles any repeated keys safely
  select(-attributes)  # drop original if you like

# Inspect
glimpse(attrs_wide)


attrs_wide$width <- attrs_wide$end - attrs_wide$start

dups <- attrs_wide$gene_name[duplicated(attrs_wide$gene_name)]

final_annot <- attrs_wide[!attrs_wide$gene_name %in% dups,]

#add weights
gw <- read.table("/Users/sfurlan/Fred Hutch Cancer Center/Furlan_Lab - General/datasets/fetalBM/data/geneWeights.tsv", header=T)
final_annot$weight <- gw$weight[match(final_annot$gene_id, gw$geneID)]
final_annot$weight[is.na(final_annot$weight)]<- 1.0
 
write.table(data.frame(ens=final_annot$gene_id, symbol=final_annot$gene_name, length=final_annot$width, weight = final_annot$weight), "/Users/sfurlan/Fred Hutch Cancer Center/Furlan_Lab - General/datasets/fetalBM/data/geneWeights_withSymbol.tsv", quote = F, sep = "\t")


```


## Run the python notebook to deconvolute

#Python notebook working as evidenced by this heatmap.
```{r}
library(ComplexHeatmap)
d <- read.csv("/Users/sfurlan/Desktop/test_fit.csv")
rownames(d)<-d$True
d$True <- NULL
d <- d[,!(grepl("LSC", colnames(d)) | grepl("rHSC", colnames(d)) | grepl("Blast", colnames(d)))]
m <- as.matrix(d)
cols <- pals::glasbey(13)
names(cols) <- unique(strsplit(colnames(m), "\\.") %>% sapply("[[", 2))
cola <- HeatmapAnnotation(bulk = strsplit(colnames(m), "\\.") %>% sapply("[[", 2), col = list(bulk=cols))
Heatmap(m, bottom_annotation = cola, show_column_names = F, col = viridis_light_high)
```

#### Claude

```{r}
#' Deconvolve bulk RNA-seq data using single-cell signatures
#'
#' @param signatures Matrix of cellular signatures (genes × cell_types). 
#'   Each column should be normalized to sum to 1.
#' @param bulk_counts Matrix of bulk RNA-seq counts (genes × samples).
#'   Gene names must match signatures.
#' @param gene_lengths Numeric vector of gene lengths, same length as nrow(signatures).
#' @param gene_weights Optional numeric vector of gene weights (0-1), same length as 
#'   nrow(signatures). Default is all 1s.
#' @param backend Which compute backend to use: "ndarray" (CPU, default), 
#'   "wgpu" (GPU if compiled with wgpu feature), "cuda" (CUDA if available).
#' @param insert_size Insert size for RNA-seq fragments. Default 500.
#' @param init_log_exp Initial value for log exposures. Default -10.
#' @param learning_rate Learning rate for optimization. Default 0.01.
#' @param l1_lambda L1 regularization parameter. Default 0.0 (no regularization).
#' @param l2_lambda L2 regularization parameter. Default 0.0 (no regularization).
#' @param max_iter Maximum number of iterations. Default 10000.
#' @param poll_interval Check convergence every N iterations. Default 100.
#' @param ll_tolerance Log-likelihood convergence tolerance. Default 1e-6.
#' @param sparsity_tolerance Sparsity convergence tolerance. Default 1e-4.
#'
#' @return A list with:
#'   \item{exposures}{Matrix of cell type proportions (cell_types+intercept × samples)}
#'   \item{pred_counts}{Matrix of predicted bulk counts (genes × samples)}
#'   \item{proportions}{Normalized cell type proportions excluding intercept}
#'
#' @export
#' @examples
#' \dontrun{
#' # CPU backend (default)
#' result <- deconvolve_bulk(
#'   signatures = signatures,
#'   bulk_counts = bulk_counts,
#'   gene_lengths = gene_lengths
#' )
#' 
#' # GPU backend (if available)
#' result <- deconvolve_bulk(
#'   signatures = signatures,
#'   bulk_counts = bulk_counts,
#'   gene_lengths = gene_lengths,
#'   backend = "wgpu"
#' )
#' }
deconvolve_bulk <- function(
    signatures,
    bulk_counts,
    gene_lengths,
    gene_weights = NULL,
    backend = "wgpu",
    insert_size = 500,
    init_log_exp = -10,
    learning_rate = 0.1,
    l1_lambda = 0.0,
    l2_lambda = 0.0,
    max_iter = 2000,
    poll_interval = 100,
    ll_tolerance = 1e-6,
    sparsity_tolerance = 1e-4
) {
    # Input validation
    if (!is.matrix(signatures)) {
        stop("signatures must be a matrix")
    }
    if (!is.matrix(bulk_counts)) {
        stop("bulk_counts must be a matrix")
    }
    if (!is.numeric(gene_lengths)) {
        stop("gene_lengths must be numeric")
    }
    
    n_genes <- nrow(signatures)
    n_cell_types <- ncol(signatures)
    
    # Check dimensions
    if (nrow(bulk_counts) != n_genes) {
        stop("Number of genes in bulk_counts (", nrow(bulk_counts), 
             ") must match signatures (", n_genes, ")")
    }
    
    if (length(gene_lengths) != n_genes) {
        stop("Length of gene_lengths (", length(gene_lengths), 
             ") must match number of genes (", n_genes, ")")
    }
    
    # Check that signatures are normalized
    col_sums <- colSums(signatures)
    if (any(abs(col_sums - 1.0) > 1e-6)) {
        warning("Some signature columns don't sum to 1. Normalizing...")
        signatures <- sweep(signatures, 2, colSums(signatures), "/")
    }
    
    # Default gene weights
    if (is.null(gene_weights)) {
        gene_weights <- rep(1.0, n_genes)
    }
    
    if (length(gene_weights) != n_genes) {
        stop("Length of gene_weights (", length(gene_weights), 
             ") must match number of genes (", n_genes, ")")
    }
    
    # Ensure gene weights are in [0, 1]
    if (any(gene_weights < 0) || any(gene_weights > 1)) {
        stop("gene_weights must be in [0, 1]")
    }
    
    # Call Rust function
    cat("Running deconvolution...\n")
    result <- fit_deconv(
        signatures,
        bulk_counts,
        gene_lengths,
        gene_weights,
        backend,
        insert_size,
        init_log_exp,
        learning_rate,
        l1_lambda,
        l2_lambda,
        max_iter,
        poll_interval,
        ll_tolerance,
        sparsity_tolerance
    )
    
    # Add row and column names
    n_samples <- ncol(bulk_counts)
    
    # Exposures: (n_cell_types + 1) × n_samples
    rownames(result$exposures) <- c(colnames(signatures), "Intercept")
    colnames(result$exposures) <- colnames(bulk_counts)
    
    # Predicted counts: n_genes × n_samples
    rownames(result$pred_counts) <- rownames(signatures)
    colnames(result$pred_counts) <- colnames(bulk_counts)
    
    # Calculate normalized proportions (excluding intercept)
    exposures_no_intercept <- result$exposures[1:n_cell_types, , drop = FALSE]
    proportions <- sweep(exposures_no_intercept, 2, 
                        colSums(exposures_no_intercept), "/")
    
    # Add proportions to result
    result$proportions <- proportions
    
    # Calculate goodness of fit metrics
    result$metrics <- calculate_fit_metrics(
        observed = bulk_counts,
        predicted = result$pred_counts
    )
    
    cat("Done!\n")
    return(result)
}


#' Calculate goodness of fit metrics
#' @keywords internal
calculate_fit_metrics <- function(observed, predicted) {
    # Per-sample metrics
    obs_total <- colSums(observed)
    pred_total <- colSums(predicted)
    
    # Correlation per sample
    correlations <- sapply(1:ncol(observed), function(i) {
        cor(observed[, i], predicted[, i], method = "pearson")
    })
    
    # RMSE per sample
    rmse <- sapply(1:ncol(observed), function(i) {
        sqrt(mean((observed[, i] - predicted[, i])^2))
    })
    
    # Pseudo-R-squared (based on Poisson deviance)
    # Null model: predicted = mean(observed)
    null_deviance <- sum(sapply(1:ncol(observed), function(i) {
        obs <- observed[, i]
        null_pred <- mean(obs)
        2 * sum(obs * log(obs / null_pred + 1e-10) - (obs - null_pred))
    }))
    
    # Full model deviance
    full_deviance <- sum(sapply(1:ncol(observed), function(i) {
        obs <- observed[, i]
        pred <- predicted[, i]
        2 * sum(obs * log(obs / (pred + 1e-10)) - (obs - pred))
    }))
    
    pseudo_r2 <- 1 - (full_deviance / null_deviance)
    
    list(
        correlation = correlations,
        rmse = rmse,
        count_ratio = pred_total / obs_total,
        pseudo_r2 = pseudo_r2,
        mean_correlation = mean(correlations),
        mean_rmse = mean(rmse)
    )
}

#' Plot deconvolution results
#'
#' @param result Output from deconvolve_bulk()
#' @param type Type of plot: "proportions", "fit", or "both"
#' @export
plot_deconvolution <- function(result, type = "both") {
    if (!requireNamespace("ggplot2", quietly = TRUE)) {
        stop("ggplot2 is required for plotting")
    }
    
    library(ggplot2)
    
    if (type %in% c("proportions", "both")) {
        # Stacked barplot of proportions
        prop_df <- as.data.frame(result$proportions)
        prop_df$CellType <- rownames(prop_df)
        prop_long <- tidyr::pivot_longer(
            prop_df, 
            -CellType, 
            names_to = "Sample", 
            values_to = "Proportion"
        )
        
        p1 <- ggplot(prop_long, aes(x = Sample, y = Proportion, fill = CellType)) +
            geom_bar(stat = "identity") +
            theme_minimal() +
            theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
            labs(title = "Cell Type Proportions", y = "Proportion", x = "Sample")
        
        print(p1)
    }
    
    if (type %in% c("fit", "both")) {
        # Observed vs predicted scatter
        obs_flat <- as.vector(result$pred_counts)
        pred_flat <- as.vector(result$pred_counts)
        
        fit_df <- data.frame(
            Observed = obs_flat,
            Predicted = pred_flat
        )
        
        p2 <- ggplot(fit_df, aes(x = log1p(Observed), y = log1p(Predicted))) +
            geom_point(alpha = 0.3) +
            geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
            theme_minimal() +
            labs(
                title = "Model Fit Quality",
                x = "log1p(Observed counts)",
                y = "log1p(Predicted counts)"
            )
        
        print(p2)
    }
}


library(Seurat)

# Load single-cell data
sc_data <- readRDS("/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchCancerCenter/Furlan_Lab - General/datasets/Healthy_BM_greenleaf/230329_rnaAugmented_seurat.RDS")

# Extract normalized counts and cell type labels
counts <- GetAssayData(sc_data, layer = "data")  # log-normalized
cell_types <- sc_data$SFClassification

# Aggregate by cell type and create signatures
signatures <- sapply(unique(cell_types), function(ct) {
  cells <- which(cell_types == ct)
  rowMeans(expm1(counts[, cells]))  # Convert back from log
})

# Normalize to proportions (sum to 1)
signatures <- sweep(signatures, 2, colSums(signatures), "/")

# Load bulk counts and gene lengths
bulk_counts <- "/Users/sfurlan/Library/CloudStorage/OneDrive-FredHutchCancerCenter/computation/Analysis/ddata/MPAL/data/bulk/GSE74246_RNAseq_All_Counts.txt"
bulk_counts <- read.csv(bulk_counts, sep = "\t")
rn <- bulk_counts$X_TranscriptID
bulk_counts$X_TranscriptID<- NULL
bulk_counts <- as.matrix(bulk_counts)
rownames(bulk_counts) <- rn
gene_info <- read.table("/Users/sfurlan/Fred Hutch Cancer Center/Furlan_Lab - General/datasets/fetalBM/data/geneWeights_withSymbol.tsv", header = TRUE)
rownames(gene_info) <- gene_info$symbol

# Ensure same genes
common_genes <- intersect(intersect(rownames(signatures), rownames(bulk_counts)), gene_info$symbol)
signatures <- signatures[common_genes, ]
bulk_counts <- bulk_counts[common_genes, ]

# Get gene lengths
gene_lengths <- gene_info[common_genes, "length"]

dim(signatures)
dim(bulk_counts)
length(gene_lengths)

#fast
  # user  system elapsed 
  # 3.852   0.809   9.530 
system.time(result <- deconvolve_bulk(
  signatures = signatures,
  bulk_counts = bulk_counts+0.00001,
  gene_lengths = gene_lengths+0.00001,
  init_log_exp = -10,  # Start with higher initial values
  learning_rate = 0.1,  # Increase learning rate
  max_iter = 50000,
  backend = "wgpu"
))

#long 
 #   user  system elapsed 
 # 36.121   6.782  39.196 

system.time(result <- deconvolve_bulk(
  signatures = signatures,
  bulk_counts = bulk_counts+0.00001,
  gene_lengths = gene_lengths+0.00001,
  init_log_exp = -10,  # Start with higher initial values
  learning_rate = 0.1,  # Increase learning rate
  max_iter = 2000,
  backend = "ndarray"
))

#very very long
#    user  system elapsed 
# 199.663 162.284 319.761 
system.time(result <- deconvolve_bulk(
  signatures = signatures,
  bulk_counts = bulk_counts+0.00001,
  gene_lengths = gene_lengths+0.00001,
  init_log_exp = -10,  # Start with higher initial values
  learning_rate = 0.1,  # Increase learning rate
  max_iter = 2000,
  backend = "candle"
))

```



```{r}

# View cell type proportions
head(result$proportions)

# Goodness of fit metrics
print(result$metrics)
cat("Mean correlation:", result$metrics$mean_correlation, "\n")
cat("Pseudo R-squared:", result$metrics$pseudo_r2, "\n")

# Plot results
plot_deconvolution(result, type = "both")

```

```{r}
res <- result$exposures
d <- res
d <- d[,!(grepl("LSC", colnames(d)) | grepl("rHSC", colnames(d)) | grepl("Blast", colnames(d)))]
m <- as.matrix(d)
sm <- scale(m)
cols <- pals::glasbey(13)
names(cols) <- unique(strsplit(colnames(m), "\\.") %>% sapply("[[", 2))
cola <- HeatmapAnnotation(bulk = strsplit(colnames(m), "\\.") %>% sapply("[[", 2), col = list(bulk=cols))
Heatmap(m, top_annotation = cola, show_column_names = F, col = viridis_light_high)
Heatmap(sm, top_annotation = cola, show_column_names = F, col = viridis_light_high, cluster_rows = F)
```

#Redo this in R

```{r}
library(viewmastR)


seuHR <- readRDS("/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchCancerCenter/Furlan_Lab - General/datasets/Healthy_BM_greenleaf/230329_rnaAugmented_seurat.RDS")
DimPlot(seuHR, group.by = "SFClassification")
bulk_count_data = "/Users/sfurlan/Library/CloudStorage/OneDrive-FredHutchCancerCenter/computation/Analysis/ddata/MPAL/data/bulk/GSE74246_RNAseq_All_Counts.txt"
gene_weights = "/Users/sfurlan/Fred Hutch Cancer Center/Furlan_Lab - General/datasets/fetalBM/data/geneWeights_withSymbol.tsv"
fit_csv = "/Users/sfurlan/Desktop/test_fit_rust.csv"


bulk_count_data <- read.csv(bulk_count_data, sep = "\t")
rn <- bulk_count_data$X_TranscriptID
bulk_count_data$X_TranscriptID<- NULL
bulk_count_data <- as.matrix(bulk_count_data)
rownames(bulk_count_data) <- rn
gene_weights <- read.table(gene_weights, sep = "\t", header = T)

pseudobulk <- function(seu, by, assay = "RNA", layer = "data"){
  meta <- seu@meta.data
  counts_mat <- GetAssayData(seu, assay = assay, layer = layer)
  # extract your cell‐type factor
  cell_types <- as.character(meta[[by]])
  unique_types <- unique(cell_types)
  
  # pre‐allocate an empty matrix: genes × n_types
  genes       <- rownames(counts_mat)
  scSigs_mat  <- matrix(0,
                        nrow = length(genes),
                        ncol = length(unique_types),
                        dimnames = list(genes, unique_types))
  
  # fill it
  for (ct in unique_types) {
    idx <- which(cell_types == ct)
    # sum counts for that cell-type
    vec <- rowSums(counts_mat[, idx, drop=FALSE])
    # normalize so they sum to 1
    scSigs_mat[, ct] <- vec / sum(vec)
  }
  scSigs_mat
}


undebug(pseudobulk)
# 1) build healthy reference from your non-malignant clusters
ref_mat <- pseudobulk(seuHR, "SFClassification" )
bulk_mat <- bulk_count_data
C_mat <- gene_weights$length
# C_mat <- bulk_count_data$size
names(C_mat)<-gene_weights$symbol
# bulk_mat <- as.matrix(bulk_count_data[,3:ncol(bulk_count_data)])
# rownames(bulk_mat)<-bulk_count_data$X

common_feats <- intersect(intersect(rownames(ref_mat), names(C_mat)), rownames(bulk_mat))


ref_mat <- ref_mat[common_feats, ]
C_mat <- C_mat[common_feats]
bulk_mat <- bulk_mat[common_feats,]
w_vec <- gene_weights$weight[match(common_feats, gene_weights$symbol)]

bulk_mat_t <- bulk_mat
dim(bulk_mat_t)

saveRDS(list(ref_mat, bulk_mat_t, C_mat, w_vec), file.path("~/Desktop/seuHR/test_list.RDS"))



library(viewmastR)

l<-readRDS("~/Desktop/seuHR/test_list.RDS")
ref_mat = l[[1]]
bulk_mat_t = l[[2]]
C_mat = l[[3]]
w_vec <- l[[4]]


result <- res$exposures
dim(result)
rownames(result) <- c(colnames(ref_mat), "Intercept")
colnames(result) <- colnames(bulk_mat_t)
d <- result
d <- d[,!(grepl("LSC", colnames(d)) | grepl("rHSC", colnames(d)) | grepl("Blast", colnames(d)))]
m <- as.matrix(d)
cols <- pals::glasbey(13)
names(cols) <- unique(strsplit(colnames(m), "\\.") %>% sapply("[[", 2))
cola <- HeatmapAnnotation(bulk = strsplit(colnames(m), "\\.") %>% sapply("[[", 2), col = list(bulk=cols))
Heatmap(m, bottom_annotation = cola, show_column_names = F, col = viridis_light_high)




```

```{r}

cds <- readRDS(file.path(ROOT_DIR4, "healthy_bm_greenleaf.rds"))
plot_cells(cds, color_cells_by = "BioClassification")

rownames(cds)<-rowData(cds)$gene_id
#cds <- normalized_counts(cds)
PCs <- 25 #Number of PCs for clustering
n_top <- c(3000,2500,2500) #number of features
cds<-iterative_LSI(cds, num_dim = PCs, num_features = n_top, resolution = c(2e-4, 9e-4, 1e-3), verbose = T)
cds<-reduce_dimension(cds, reduction_method = "UMAP", preprocess_method = "LSI", umap.min_dist = 0.45,  umap.n_neighbors = 35, umap.metric = "euclidean", verbose=T)
plot_cells(cds, color_cells_by = "BioClassification")


feats <- cds@int_metadata$LSI_model$features

#make a seurat object for export
counts <- cds@assays@data$counts
rownames(counts)<-rowData(cds)$gene_id
seuHR <- CreateSeuratObject(counts = counts , meta.data = data.frame(colData(cds)))
seuHR <- NormalizeData(seuHR) %>% ScaleData()
lsimat <- cds@int_metadata$LSI_model$svd$v
rownames(lsimat)<-colnames(cds)
colnames(lsimat)<-paste0("LSI_", 1:ncol(lsimat))
seuHR$lsi <- CreateDimReducObject(embeddings = lsimat, assay = "RNA")
# seuHR <- readRDS(file.path(ROOT_DIR4, "healthy_bm_greenleaf.rds"))
# seuHR <- CreateSeuratObject(counts = seuHR@assays@data$counts, meta.data = data.frame(colData(seuHR)))
# seuHR <- NormalizeData(seuHR) %>% ScaleData()
# VariableFeatures(seuHR)<- feats
# seuHR <- Signac::RunTFIDF(seuHR)
# seuHR <- Signac::RunSVD(seuHR)
VariableFeatures(seuHR)<-feats
seuHR <- RunUMAP(seuHR, dims = 1:25, reduction = "lsi", min.dist = 0.45, n.neighbors = 35, metric = "euclidean")
DimPlot(seuHR, group.by = "BioClassification")
DimPlot(seuHR, group.by = "Group")

dir.create("~/Desktop/seuHR")
flscuts::export_seurat_to_10x(seuHR, dir = "~/Desktop/seuHR")
```




```{r}

library(viewmastR)

bulk_count_data = "/Users/sfurlan/Fred Hutch Cancer Center/Furlan_Lab - General/datasets/AML/1031/coembed/data/aml_all_bulk_counts.csv"
gene_weights = "/Users/sfurlan/Fred Hutch Cancer Center/Furlan_Lab - General/datasets/fetalBM/data/geneWeights.tsv"

bulk_count_data <- read.csv(bulk_count_data)
gene_weights <- read.table(gene_weights, sep = "\t", header = T)

pseudobulk <- function(seu, by, assay = "RNA", layer = "counts"){
  meta <- seu@meta.data
  counts_mat <- GetAssayData(seu, assay = assay, layer = layer)
  # extract your cell‐type factor
  cell_types <- as.character(meta[[by]])
  unique_types <- unique(cell_types)
  
  # pre‐allocate an empty matrix: genes × n_types
  genes       <- rownames(counts_mat)
  scSigs_mat  <- matrix(0,
                        nrow = length(genes),
                        ncol = length(unique_types),
                        dimnames = list(genes, unique_types))
  
  # fill it
  for (ct in unique_types) {
    idx <- which(cell_types == ct)
    # sum counts for that cell-type
    vec <- rowSums(counts_mat[, idx, drop=FALSE])
    # normalize so they sum to 1
    scSigs_mat[, ct] <- vec / sum(vec)
  }
  scSigs_mat
}



# 1) build healthy reference from your non-malignant clusters
ref_mat <- pseudobulk(seuHR, "BioClassification" )
C_mat <- bulk_count_data$size
names(C_mat)<-bulk_count_data$X
bulk_mat <- as.matrix(bulk_count_data[,3:ncol(bulk_count_data)])
rownames(bulk_mat)<-bulk_count_data$X

common_feats <- intersect(intersect(rownames(ref_mat), names(C_mat)), gene_weights$geneID)


ref_mat <- ref_mat[common_feats, ]
C_mat <- C_mat[common_feats]
bulk_mat <- bulk_mat[common_feats,]
w_vec <- gene_weights$weight[match(common_feats, gene_weights$geneID)]


tester <- c(2116, 3432, 3391, 963, 4374, 1277, 3943, 1295, 3714, 3576, 1263, 2275, 583, 2497, 889, 1380, 4483, 4431, 3146, 1955, 1997, 133, 3573, 3622, 660, 2902, 30, 1017, 2783, 1919, 2996, 4908, 2051, 1474, 1018, 631, 180, 3758, 2229, 986, 2302, 2197, 3451, 3267, 2347, 4867, 2328, 2638, 2827, 737, 371, 4592, 647, 3349, 3562, 3482, 2389, 489, 3046, 479, 2293, 1079, 2131, 4072, 1907, 3456, 2255, 540, 4294, 3011, 4610, 2970, 4819, 3088, 3114, 584, 192, 2751, 945, 1158, 1696, 3889, 1020, 2533, 1579, 4598, 3182, 2362, 1580, 2713, 3603, 4635, 3678, 2451, 2332, 4, 2168, 3370, 3502, 305)

bulk_mat_t <- bulk_mat[,tester]
dim(bulk_mat_t)

#saveRDS(list(ref_mat, bulk_mat_t, C_mat, w_vec), file.path("~/Desktop/seuHR/test_list.RDS"))



library(viewmastR)

l<-readRDS("~/Desktop/seuHR/test_list.RDS")
ref_mat = l[[1]]
bulk_mat_t = l[[2]]
C_mat = l[[3]]
w_vec <- l[[4]]

deconvolute <- function(
    ref_mat, 
    bulk_mat, 
    length_vec, 
    weight_vec,
    init_log_exp = -5,
    lr           = 1e-3,
    l1_lambda    = 0.01,
    l2_lambda    = 0.01,
    epochs       = 300)
{
    fit <- fit_deconv(
      sigs = ref_mat,
      bulk = bulk_mat,
      mol2read = length_vec,
      w_vec = weight_vec,
      init_log_exp =init_log_exp,
      lr          = lr,
      l1_lambda   = l1_lambda,
      l2_lambda   = l2_lambda,
      epochs      = epochs)
  
        ## -- 1. sort the sample columns the way pandas .sort_index(axis=1) does ----
      loss <- fit$loss
      fit <- fit$exposures[ , order(colnames(bulk_mat))]
      # 
      # 
      # ## -- 2. split out “goodness-of-fit” rows -----------------------------------
      # gof_noms <- c("pR2", "fitCount", "obsCount")
      # gof <- fit[gof_noms, , drop = FALSE]
      # 
      # ## add log2(countRatio) = log2(fitCount / obsCount) per sample
      # gof["log2(countRatio)", ] <- log2(gof["fitCount", ] / gof["obsCount", ])
      # 
      # ## -- 3. raw exposures (everything that is *not* a GOF row) -----------------
      # expos_idx <- !rownames(fit) %in% gof_noms
      # raw <- fit[expos_idx, , drop = FALSE]
    
      ## -- 4. column-wise normalisation so each sample sums to 1 -----------------
      exposures <- sweep(fit, 2, colSums(fit), FUN = "/")
    
      ## -- 5. return a list mimicking the Python dict ----------------------------
      list(
        exposures = exposures
        # gof       = gof,
        # raw       = raw          # raw (unnormalised) exposure matrix
      )
}

# C_mat <- matrix(
#   as.numeric(C_mat),
#   nrow = length(C_mat),
#   ncol = ncol(bulk_mat_t),
#   byrow = FALSE
# )

dim(ref_mat)
length(C_mat)
dim(bulk_mat_t)
length(w_vec)

undebug(deconvolute)
res <- deconvolute(ref_mat, bulk_mat_t, length_vec = C_mat, weight_vec = w_vec, lr = 0.01, epochs = 3000)


ComplexHeatmap::Heatmap(res$exposures, cluster_rows = F, cluster_columns = F,col = viridis_light_high)
```

```{r}
#debug(flscuts::iterative_LSI)


cor(cds@int_metadata$LSI_model$svd$d, seuHR@reductions$lsi@misc$svd$d)       # ≈ 1.00

identical(
  cds@int_metadata$LSI_model$features,
  seuHR@reductions$lsi@misc$features
)

identical(
  cds@int_metadata$LSI_model$features,
  VariableFeatures(seuHR)
)

library(SeuratDisk)
SeuratDisk::SaveH5Seurat(seuHR, filename = file.path("~/Desktop/seuHR.h5Seurat"))
SeuratDisk::Convert("~/Desktop/seuHR.h5Seurat", dest = "h5ad", overwrite = T)


```




```{r}
filter_calls <- function(seu_obj, training_result, prob_thresh = 0.95, pop_thresh = 0.05, labels, output_name = "tumor_call"){
  max <- apply(training_result$training_output$probs, 1, max)
  names(max) <- labels[apply(training_result$training_output$probs, 1, which.max)]
  names(max)[max<prob_thresh]<- NA
  seu_obj[[output_name]] <- names(max)
  tk <- names(table(names(max)))[table(names(max)) > pop_thresh * sum(table(names(max)))]
  seu_obj[[paste0(output_name, "_filtered")]]<- NA
  seu_obj[[paste0(output_name, "_filtered")]][names(max) %in% tk,1] <- names(max)[names(max) %in% tk]
  return(seu_obj)
}


```



```{r}
plot_gene_dispersion(seuHR)
seuHR <- select_features(seuHR, top_n = 5000, logmean_ul = -1, logmean_ll = -9)

vgHR <- get_selected_features(seuHR)

seuF <- calculate_feature_dispersion(seuF)
plot_gene_dispersion(seuF)
seuF <- select_features(seuF, top_n = 5000, logmean_ul = -3, logmean_ll = -10)
vgF <- get_selected_features(seuF)

seuF <- NormalizeData(seuF) %>% ScaleData()


res <- viewmastR(query_cds = seuHR, ref_cds = seuF, ref_celldata_col = "cell.labels", selected_features = vgF, return_probs = T, return_type = "list")
seuHR <- filter_calls(seuHR, prob_thresh = .9, pop_thresh = 0.001, training_result = res, labels = levels(factor(seuF$cell.labels)), output_name = "fetal")

DimPlot(seuHR, group.by = "fetal_filtered", cols = as.character(c(pals::alphabet(), pals::polychrome(), pals::glasbey())))


seu3 <- readRDS('/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchCancerCenter/Furlan_Lab - General/experiments/Peds_UCSF/cds/KMT2Apatients_seuratobject_cleaned.rds')

res3 <- viewmastR(query_cds = seu3, ref_cds = seuF, ref_celldata_col = "cell.labels", selected_features = vgF, return_probs = T, return_type = "list")
seu3 <- filter_calls(seu3, prob_thresh = .2, pop_thresh = 0.001, training_result = res3, labels = levels(factor(seuF$cell.labels)), output_name = "fetal")
DimPlot(seu3, group.by = "fetal_filtered", cols = as.character(c(pals::alphabet(), pals::polychrome(), pals::glasbey())))


plotly::ggplotly(DimPlot(seuHR, group.by = "viewmastR_pred", cols = as.character(c(pals::alphabet(), pals::polychrome(), pals::glasbey()))))

seuF <- viewmastR(seuF, seuHR, ref_celldata_col = "SFClassification", selected_features = vgHR)
seuF@meta.data
DimPlot(seuF, group.by = "viewmastR_pred", cols = cols)
FeaturePlot_scCustom(seuF, features = "FOXP3")
plotly::ggplotly(DimPlot(seuF, group.by = "cell.labels", raster = F, cols = as.character(c(pals::alphabet(), pals::polychrome(), pals::glasbey()))))
```


```{r}
colnames(seuHR@assays$RNA@data)

n_top <- c(3000,2500,2500) #number of features
seuHR <- flscuts::iterative_LSI(seuHR, num_dim = 25, num_features = n_top, resolution = c(2e-4, 9e-4, 1e-3), verbose = T, do_tf_idf = F)


undebug(flscuts::iterative_LSI)


single_cell_ref_column = "cell.labels"
bulk_count_data = file.path(ROOT_DIR3, "aml_all_bulk_counts.csv")
# bulk_meta_data = "~/Fred Hutchinson Cancer Research Center/Furlan_Lab - General/datasets/AML/1031/coembed/data/bulk_meta.csv" #don't really need this
gene_weights = file.path(ROOT_DIR2, "geneWeights.tsv")


seuHR <- file.path()
```

```{r}
library(Seurat)
seu <- readRDS(single_cell_ref)
plotly::ggplotly(DimPlot(seu, group.by = "cell.labels", raster = F, cols = as.character(c(pals::alphabet(), pals::polychrome(), pals::glasbey()))))

```



## Running viewmastR

```{r, dpi=300, fig.height=4, fig.width = 6}
suppressPackageStartupMessages({
library(viewmastR)
library(Seurat)
library(ggplot2)
})

#query dataset
seu <- readRDS(file.path(ROOT_DIR1, "240813_final_object.RDS"))
#reference dataset
seur<-readRDS(file.path(ROOT_DIR2, "230329_rnaAugmented_seurat.RDS"))
vg <- get_selected_features(seu)
```

## Build the model and infer for small dataset (not using chunks and parallelization)

This is also covered elsewhere
```{r}
results <- viewmastR(seu, seur, ref_celldata_col = "SFClassification", selected_features = vg, max_epochs = 4, train_only = T)



seu<-viewmastR_infer(seu, results[["model_dir"]], vg, labels = levels(factor(seur$SFClassification)), batch_size = 250)
DimPlot(seu, group.by = "viewmastR_inferred", cols = seur@misc$colors)
```

## Build the model and infer for large dataset (dividing the query into chunks and using parallelization)

By using chunks and workers, you can infer from the model only chunks at a time using multiple workers in parallel.
```{r}
options(future.globals.maxSize = 16 * 1024^3)
seu<-viewmastR_infer(seu, results[["model_dir"]], query_celldata_col = "viewmastR_inferred_parallel", vg, labels = levels(factor(seur$SFClassification)), chunks = 4, workers = 4)
DimPlot(seu, group.by = "viewmastR_inferred_parallel", cols = seur@misc$colors)
```

We see no difference if parallelization is used
```{r, dpi=300, fig.height=6, fig.width=10}
confusion_matrix(pred = factor(seu$viewmastR_inferred), gt = factor(seu$viewmastR_inferred_parallel), cols = seur@misc$colors)
```


## Bigger dataset
By optimising chunk size, workers and batch size, you can achieve faster results on big datasets. Here we are just inferring back on the larger reference set (because it is larger, but this is not advisable in normal practice)
```{r}
seur<-viewmastR_infer(seur, results[["model_dir"]], query_celldata_col = "viewmastR_inferred_parallel", vg, labels = levels(factor(seur$SFClassification)), chunks = 20, workers = 16, batch_size = 3000)
DimPlot(seur, group.by = "viewmastR_inferred_parallel", cols = seur@misc$colors)
```

```{r, dpi=300, fig.height=6, fig.width=10}
confusion_matrix(pred = factor(seur$SFClassification), gt = factor(seur$viewmastR_inferred_parallel), cols = seur@misc$colors)
```

