---
title: "How to augment data using viewmastR"
output: html_document
date: "2024-01-23"
always_allow_html: true
editor_options: 
  
  chunk_output_type: console
---

## Installing Rust

First you need to have an updated Rust installation. Go to this [site](https://www.rust-lang.org/tools/install) to learn how to install Rust.


## Installing viewmastR

You will need to have the devtools package installed...

```{r, eval=F}
devtools::install_github("furlan-lab/viewmastR")
```

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
rm(list=ls())
if(grepl("^gizmo", Sys.info()["nodename"])){
  ROOT_DIR1<-"/fh/fast/furlan_s/experiments/MB_10X_5p/cds"
  ROOT_DIR2<-"/fh/fast/furlan_s/grp/data/ddata/BM_data"
  ROOT_DIR3<-"/fh/fast/furlan_s/experiments/LRPBMC/cds"
} else {
  ROOT_DIR1<-"/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchinsonCancerCenter/Furlan_Lab - General/experiments/MB_10X_5p/cds"
  ROOT_DIR2<-"/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchinsonCancerCenter/Furlan_Lab - General/datasets/Healthy_BM_greenleaf"
  ROOT_DIR3<-"/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchinsonCancerCenter/Furlan_Lab - General/experiments/LRPBMC/cds"
}
```

## Load a dataset

```{r, dpi=300, fig.height=4, fig.width = 6, warning=F, message=F}
suppressPackageStartupMessages({
library(viewmastR)
library(Seurat)
library(ggplot2)
library(scCustomize)
library(magrittr)
})


#query dataset
seu<-readRDS(file.path(ROOT_DIR1, "220302_final_object.RDS"))

```

## Let's first fix this dataset and label with Tregs 

Tregs are located in cluster 9 (as evidenced by FOXP3 expression), but don't have the correct label.  Let's fix that.
```{r, dpi=300, fig.height=4, fig.width = 6}
DimPlot_scCustom(seu, group.by = "celltype")
Idents(seu)<-seu$celltype
Cluster_Highlight_Plot(seu, cluster_name = "Treg")
DimPlot_scCustom(seu, group.by = "seurat_clusters")
FeaturePlot_scCustom(seu, features = "FOXP3")
```

```{r}
seu$celltype[seu$seurat_clusters %in% "9"]<-"Treg"
DimPlot_scCustom(seu, group.by = "celltype")
Idents(seu)<-seu$celltype
Cluster_Highlight_Plot(seu, cluster_name = "Treg")
table(seu$celltype)
```

Better...

## Let's augment rare celltypes for better learning but bringing any celltype to a minimum of 1500 cells.  use the prune option to downsample any celltypes present above the norm_number down to that number.

```{r}

seuA<-augment_data(seu, "celltype", norm_number = 500, prune = T)
seuA <- FindVariableFeatures(seuA, selection.method = "vst", nfeatures = 3000) %>% NormalizeData() %>% ScaleData()
seuA <- RunPCA(seuA, features = VariableFeatures(object = seuA), npcs = 50)
ElbowPlot(seuA, 50) 
seuA<- FindNeighbors(seuA, dims = 1:40) %>% FindClusters(resolution = 2) %>% RunUMAP(dims = 1:40, n.components = 2)
DimPlot_scCustom(seuA, group.by = "celltype")
FeaturePlot_scCustom(seuA, features = "FOXP3")

```


## Now we train another PBMC dataset with labels for tregs that are questionable and see if we can find them
```{r}

seu<-readRDS(file.path(ROOT_DIR3, "240919_5p_seu.RDS"))

seu<-calculate_gene_dispersion(seu)
seu<-select_features(seu, top_n = 10000, logmean_ul = -1, logmean_ll = -8)
vgq<-get_selected_features(seu)

seuA<-calculate_gene_dispersion(seuA)
seuA<-select_features(seuA, top_n = 10000, logmean_ul = -1, logmean_ll = -8)
vgr<-get_selected_features(seuA)

vg<-intersect(vgq, vgr)

seu<-viewmastR(seu, seuA, ref_celldata_col = "celltype", selected_features = vg)

DimPlot_scCustom(seu, group.by = "mCelltype")
#need to fix spelling error
#seu$mCelltype[seu$mCelltype=="Jeutrophils"]<-"Neutrophils"
DimPlot_scCustom(seu, group.by = "viewmastR_pred")
```


## A confusion matrix showing prediction vs ground truth
```{r, dpi=300, fig.height=6, fig.width=10}
co<-as.character(pals::polychrome(length(levels(factor(c(seu$mCelltype, seu$viewmastR_pred))))))
names(co)<-levels(factor(c(seu$mCelltype, seu$viewmastR_pred)))
confusion_matrix(pred = factor(seu$viewmastR_pred), gt = factor(seu$mCelltype), cols = co)
Idents(seu)<-seu$viewmastR_pred
VlnPlot_scCustom(seu, features = "FOXP3")
Idents(seu)<-seu$mCelltype
VlnPlot_scCustom(seu, features = "FOXP3")
```



## Appendix
```{r Appendix}
sessionInfo()
getwd()
```
