---
title: "How to use viewmastR with large query objects"
output: html_document
date: "2024-12-10"
always_allow_html: true
editor_options: 
  chunk_output_type: console
---


```{r, echo=FALSE, message=FALSE, warning=FALSE}
rm(list=ls())
if(grepl("^gizmo", Sys.info()["nodename"])){
  ROOT_DIR1<-"/fh/fast/furlan_s/experiments/MB_10X_5p/cds"
  ROOT_DIR2<-"/fh/fast/furlan_s/grp/data/ddata/BM_data"
} else {
  ROOT_DIR1<-"/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchinsonCancerCenter/Furlan_Lab - General/experiments/MB_10X_5p/cds"
  ROOT_DIR2<-"/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchinsonCancerCenter/Furlan_Lab - General/datasets/Healthy_BM_greenleaf"
}
```


## Installing Rust

First you need to have an updated Rust installation. Go to this [site](https://www.rust-lang.org/tools/install) to learn how to install Rust.


## Installing viewmastR

You will need to have the devtools package installed...

```{r, eval=F}
devtools::install_github("furlan-lab/viewmastR")
```


## Running viewmastR

```{r, dpi=300, fig.height=4, fig.width = 6}
suppressPackageStartupMessages({
  library(viewmastR)
  library(Seurat)
  library(ggplot2)
})

#query dataset
#seu <- readRDS(file.path(ROOT_DIR1, "240813_final_object.RDS"))
seu <- qs2::qs_read("/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchCancerCenter/Furlan_Lab_2025 - General/experiments/AML_MRD/DL2_SF/cds/260113_Filtered_Processed_vmAnnotated_scrublet.qs")
#reference dataset
seur<-readRDS(file.path(ROOT_DIR2, "230329_rnaAugmented_seurat.RDS"))
vg <- get_selected_features(seu)
vg <- vg[vg %in% rownames(seur)]


```

## Build the model and infer for small dataset (not using chunks and parallelization)

This is also covered elsewhere
```{r}
results <- viewmastR(seu, seur, ref_celldata_col = "SFClassification", selected_features = vg, max_epochs = 4, train_only = T)



seu<-viewmastR_infer(seu, results[["model_dir"]], vg, labels = levels(factor(seur$SFClassification)))


DimPlot(seu, group.by = "viewmastR_inferred", cols = seur@misc$colors)
```

## Build the model and infer for large dataset (dividing the query into chunks and using parallelization)

By using chunks and workers, you can infer from the model only chunks at a time using multiple workers in parallel.
```{r}
gc()


system.time({
  seu <- viewmastR_infer(seu, results[["model_dir"]], 
                         query_celldata_col = "test1", vg, 
                         labels = levels(factor(seur$SFClassification)), 
                         threads = 1)
})

# Parallel nd (8 threads)
system.time({
  seu <- viewmastR_infer(seu, results[["model_dir"]], 
                         query_celldata_col = "test2", vg, 
                         labels = levels(factor(seur$SFClassification)), 
                         threads = 8)
})

all(seu$test1==seu$test2)
```

We see no difference if parallelization is used
```{r, dpi=300, fig.height=6, fig.width=10}
confusion_matrix(pred = factor(seu$test1), gt = factor(seu$test2), cols = seur@misc$colors)
```


