---
title: "How to deconvolute bulk RNAseq"
output: html_document
date: "2025-10-13"
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE, message=FALSE, warning=FALSE, include = F}
rm(list=ls())

seurat_object_loc <- "/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchCancerCenter/Furlan_Lab - General/datasets/Healthy_BM_greenleaf/230329_rnaAugmented_seurat.RDS"
```


#### Load a seurat object

This is an atlas of healthy bone marrow celltypes taken from:

Granja JM, Klemm S, McGinnis LM, Kathiria AS, Mezger A, Corces MR, Parks B, Gars E, Liedtke M, Zheng GXY, Chang HY, Majeti R, Greenleaf WJ. Single-cell multiomic analysis identifies regulatory programs in mixed-phenotype acute leukemia. Nat Biotechnol. 2019 Dec;37(12):1458-1465. doi: 10.1038/s41587-019-0332-7. Epub 2019 Dec 2. PMID: 31792411; PMCID: PMC7258684.

These data are not included in the viewmastR package.

```{r, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({library(Seurat)
  library(viewmastR)
  library(magrittr)
  library(viridis)
  library(ggplot2)
  library(ComplexHeatmap)})

# Load single-cell data
sc_data <- readRDS(seurat_object_loc)
DimPlot(sc_data, group.by = "SFClassification", cols = sc_data@misc$colors)
```

Now we pseudobulk the samples to get a matrix of "signatures"

```{r, message=FALSE, warning=FALSE}
# Extract normalized counts and cell type labels
counts <- GetAssayData(sc_data, layer = "data")  # log-normalized
cell_types <- sc_data$SFClassification

# Aggregate by cell type and create signatures
signatures <- sapply(unique(cell_types), function(ct) {
  cells <- which(cell_types == ct)
  rowMeans(expm1(counts[, cells]))  # expm1 reverses log1p normalization
})
```

## Load bulk data

Next we load data released from the same publication that has bulk RNA seq from sorted populations of bone marrow cells.  These data are included in the viewmastR package.
```{r, message=FALSE, warning=FALSE}
# Load bulk counts and gene lengths
bulk_counts <- system.file("extdata", "deconvolute", "GSE74246_RNAseq_All_Counts.txt", package = "viewmastR")
bulk_counts <- read.csv(bulk_counts, sep = "\t")
rn <- bulk_counts$X_TranscriptID
bulk_counts$X_TranscriptID<- NULL
bulk_counts <- as.matrix(bulk_counts)
rownames(bulk_counts) <- rn

gene_info <- read.table(system.file("extdata", "deconvolute", "geneWeights_withSymbol.tsv", package = "viewmastR"), header = TRUE)
rownames(gene_info) <- gene_info$symbol
```


## Feature parity

Now we make sure the features are the same across all the data used for deconvolution.  Check the dimensions to amake sure they match.  We have 21 celltypes and 81 bulk samples.  Here we find 18,879 features in common across all the inputs

```{r, message=FALSE, warning=FALSE}
# Ensure same genes
common_genes <- intersect(intersect(rownames(signatures), rownames(bulk_counts)), gene_info$symbol)
signatures <- signatures[common_genes, ]
bulk_counts <- bulk_counts[common_genes, ]

# Get gene lengths
gene_lengths <- gene_info[common_genes, "length"]
gene_weights <- gene_info[common_genes, "weight"]

dim(signatures)
dim(bulk_counts)
length(gene_lengths)
```


## Deconvolute

You can read more on our implementation below, but here is a minimally working example using default parameters.

```{r}

system.time(result_gd <- deconvolve_bulk(
  signatures = signatures,
  bulk_counts = bulk_counts,
  gene_lengths = gene_lengths,
  gene_weights = gene_weights, 
  max_iter = 3000
))

system.time(result_em <- deconvolve_bulk(
  signatures = signatures,
  bulk_counts = bulk_counts,
  gene_lengths = gene_lengths,
  gene_weights = gene_weights, 
  max_iter = 3000,
  method = "em"
))
```

## Visualize

We will remove some of the bulk data that contains tumor cells or recombinant HSC for simplicity.  We will also plot the Intercept as column annotation so one can get a sense of the degree of signal that is unaccounted for.

```{r, message=FALSE, warning=FALSE}

data <- as.matrix(result_gd$proportions_with_intercept)
data <- data[,!(grepl("LSC", colnames(data)) | grepl("rHSC", colnames(data)) | grepl("Blast", colnames(data)))]
cols <- pals::glasbey(13)
names(cols) <- unique(strsplit(colnames(data), "\\.") %>% sapply("[[", 2))
cola <- HeatmapAnnotation(bulk = strsplit(colnames(data), "\\.") %>% sapply("[[", 2), col = list(bulk=cols))
Heatmap(data[!rownames(data) %in% "Intercept",], top_annotation = cola, show_column_names = F, col = viridis(100), bottom_annotation = HeatmapAnnotation(Intercept = anno_points(data[rownames(data) %in% "Intercept",])), name = "GD Result")


data <- as.matrix(result_em$proportions_with_intercept)
data <- data[,!(grepl("LSC", colnames(data)) | grepl("rHSC", colnames(data)) | grepl("Blast", colnames(data)))]
cols <- pals::glasbey(13)
names(cols) <- unique(strsplit(colnames(data), "\\.") %>% sapply("[[", 2))
cola <- HeatmapAnnotation(bulk = strsplit(colnames(data), "\\.") %>% sapply("[[", 2), col = list(bulk=cols))
Heatmap(data[!rownames(data) %in% "Intercept",], top_annotation = cola, show_column_names = F, col = viridis(100), bottom_annotation = HeatmapAnnotation(Intercept = anno_points(data[rownames(data) %in% "Intercept",])), name = "EM Result")

```

One can also use our build in plotting functions

```{r, message=FALSE, warning=FALSE}

plot_deconvolution(result_gd)
```


```{r, message=FALSE, warning=FALSE}
df <- data.frame(GradDescent=as.vector(as.matrix(result_gd$proportions_with_intercept)), ExpMax=as.vector(result_em$proportions_with_intercept), Celltype = rep(rownames(as.matrix(result_gd$proportions_with_intercept)), ncol(as.matrix(result_gd$proportions_with_intercept))), Sample = rep(strsplit(colnames(as.matrix(result_em$proportions_with_intercept)), "\\.") %>% sapply("[[", 2), each = nrow(as.matrix(result_em$proportions_with_intercept))))
g <- ggplot(df, aes(x=GradDescent, y=ExpMax, colour = Celltype))+geom_point()+scale_color_manual(values=c(sc_data@misc$colors, "Intercept"="black"))
plotly::ggplotly(g)

g <- ggplot(df, aes(x=GradDescent, y=ExpMax, colour = Sample))+geom_point()+scale_color_manual(values=pals::glasbey(17))
plotly::ggplotly(g)


cor(df$GradDescent, df$ExpMax)



hex_plot <- function(result, bulk_counts){
  cor(as.vector(result$pred_counts[gene_weights>0,]), as.vector(bulk_counts[gene_weights>0,]))
  obs_flat <- as.vector(bulk_counts[gene_weights>0,])
  pred_flat <- as.vector(result$pred_counts[gene_weights>0,])
  # Remove zeros for log scale
  keep_idx <- obs_flat > 0 & pred_flat > 0
  fit_df <- data.frame(
    Observed = log1p(obs_flat[keep_idx]),
    Predicted = log1p(pred_flat[keep_idx])
  )
  rsq <- cor(fit_df$Observed, fit_df$Predicted)^2
ggplot(fit_df, aes(x=Observed, y=Predicted))+geom_hex(bins = 100)+scale_fill_viridis_c(trans = "log10")+ggtitle(rsq)
}



hex_plot(result_em, bulk_counts = bulk_counts)
hex_plot(result_gd, bulk_counts = bulk_counts)


library(MuSiC)
library(SingleCellExperiment)
sce <- SingleCellExperiment(list(counts = sc_data@assays$RNA@counts), colData = DataFrame(sc_data@meta.data), rowData=DataFrame(symbol=rownames(sc_data)))

sce$sample <- "sample"

prop <- music_prop(bulk_counts, sce, markers = NULL, clusters = "SFClassification", samples = "orig.ident")

data <- as.matrix(t(prop$Est.prop.weighted))
data <- data[,!(grepl("LSC", colnames(data)) | grepl("rHSC", colnames(data)) | grepl("Blast", colnames(data)))]
cols <- pals::glasbey(13)
names(cols) <- unique(strsplit(colnames(data), "\\.") %>% sapply("[[", 2))
cola <- HeatmapAnnotation(bulk = strsplit(colnames(data), "\\.") %>% sapply("[[", 2), col = list(bulk=cols))
Heatmap(data, top_annotation = cola, show_column_names = F, col = viridis(100), name = "MuSiC Result")



df <- data.frame(MuSiC=as.vector(as.matrix(t(prop$Est.prop.weighted))), ExpMax=as.vector(result_em$proportions), Celltype = rep(rownames(as.matrix(result_em$proportions)), ncol(as.matrix(result_em$proportions))), Sample = rep(strsplit(colnames(as.matrix(result_em$proportions)), "\\.") %>% sapply("[[", 2), each = nrow(as.matrix(result_em$proportions))))

g <- ggplot(df, aes(x=MuSiC, y=ExpMax, colour = Celltype))+geom_point()+scale_color_manual(values=c(sc_data@misc$colors, "Intercept"="black"))
plotly::ggplotly(g)

g <- ggplot(df, aes(x=MuSiC, y=ExpMax, colour = Sample))+geom_point()+scale_color_manual(values=pals::glasbey(17))
plotly::ggplotly(g)

cor(df$MuSiC, df$ExpMax)

df <- data.frame(MuSiC=as.vector(as.matrix(t(prop$Est.prop.weighted))), GD=as.vector(result_gd$proportions), Celltype = rep(rownames(as.matrix(result_gd$proportions)), ncol(as.matrix(result_gd$proportions))), Sample = rep(strsplit(colnames(as.matrix(result_em$proportions)), "\\.") %>% sapply("[[", 2), each = nrow(as.matrix(result_em$proportions))))
g <- ggplot(df, aes(x=MuSiC, y=GD, colour = Celltype))+geom_point()+scale_color_manual(values=c(sc_data@misc$colors, "Intercept"="black"))
plotly::ggplotly(g)

g <- ggplot(df, aes(x=MuSiC, y=GD, colour = Sample))+geom_point()+scale_color_manual(values=pals::glasbey(17))
plotly::ggplotly(g)

cor(df$MuSiC, df$GD)
```


```{r, include = F, eval = F}
# in progress

# # # Extract tumor program
# tumor_prog <- extract_tumor_program(result_em, signatures, bulk_counts,
#                                      gene_names = rownames(signatures),
#                                      top_n = 500)
# # 
# # # View top tumor genes
# head(tumor_prog$tumor_upregulated, 20)
# head(tumor_prog$tumor_downregulated, 20)
# # 
# # # Visualize
# visualize_tumor_program(tumor_prog, bulk)
# # 
# # # Check consistency
#  consistency <- compare_tumor_consistency(tumor_prog)
#  length(consistency$consistent_upregulated)  # Core tumor program
# # 
# # # Export for pathway analysis
# export_tumor_genes(tumor_prog, "AML_tumor_program")

```

# Mathematical Summary of Poisson regression with gradient descent

**Inputs:** Signatures `S` (G×C), bulk counts `k` (G×N), gene lengths `L`, weights `w`  
**Output:** Exposures `E` (C×N) = cell type abundances

---

## Model

### Forward Pass
```
q = [S | 1/G] × [E; intercept]              # Molecules
ŷ[g,n] = q[g,n] × (L[g] / insert_size)     # Reads
k[g,n] ~ Poisson(ŷ[g,n])                    # Observed counts
```

### Parameterization
```
E = exp(log_E),  intercept = exp(log_intercept)
log_E[c,n] ~ Normal(-5, 0.1)  # Init breaks symmetry
```

### Loss Function
```
NLL = Σ w[g] × [k·log(k) - k - k·log(ŷ+ε) + ŷ]  # Poisson likelihood
L₁  = λ₁ × (NLL/100) × Σ E                      # Sparsity penalty
L₂  = λ₂ × (NLL/100) × Σ E²                     # Stability penalty
Loss = NLL + L₁ + L₂
```

---

## Optimization

**Two phases:**
1. **NULL:** Freeze E≈0, optimize intercept only → baseline
2. **Full:** Re-init E, optimize all parameters with Adam (lr=0.01)

**Convergence:** Stop when Δloss/loss < 10⁻⁶ AND Δsparsity < 10⁻⁴

---

## Output

```
Exposures:   E[c,n] = exp(log_E[c,n])           # Cell-equivalents
Proportions: P[c,n] = E[c,n] / Σ_c E[c,n]       # Fractions (sum to 1)
Quality:     intercept / (Σ E + intercept)       # <0.1 = good fit
```

---

## Key Parameters

| Parameter | Default | Range | Purpose |
|-----------|---------|-------|---------|
| `l1_lambda` | 0 | 0-10 | Sparsity (1=moderate, 5=strong) |
| `l2_lambda` | 0 | 0-10 | Stability (prevents extreme values) |
| `learning_rate` | 0.01 | 0.001-0.1 | Optimization step size |
| `insert_size` | 500 | 150-500 | Fragment length correction |

**Recommendation:** Start with defaults (no regularization). Add `l1_lambda=0.5-2` only if too many cell types.

---

## Properties

- **Non-negativity:** Guaranteed by exp() transform
- **Scale:** Set by signature normalization (Σ S[g,c] = 1)
- **Gene length correction:** Essential for RNA-seq (longer genes → more reads)
- **Performance:** 10-30 sec GPU, 1-5 min CPU (typical: 18k genes, 21 types, 81 samples)

# Mathematical Summary of Expectation-Maximization for Deconvolution

**Inputs:** Signatures `S` (G×C), bulk counts `k` (G×N), gene lengths `L`, weights `w`  
**Output:** Exposures `E` (C×N) = cell type abundances

---

## Probabilistic Model

### Generative Process
```
For each gene g in sample n:
  1. Each read comes from source j ∈ {cell types, noise}
     P(source=j) ∝ S[g,j] × E[j,n]
  
  2. Noise has uniform signature: S[g,noise] = 1/G
  
  3. Observed counts: k[g,n] ~ Poisson(ŷ[g,n])
     where ŷ[g,n] = Σ_j S[g,j] × E[j,n] × (L[g] / insert_size)
```

### Hidden Variable
```
z[g,j,n] = # of reads from gene g, sample n, assigned to source j
           (unobserved, inferred by EM)
```

---

## EM Algorithm

### E-step: Compute Responsibilities
```
For each source j (cell type or noise):
  contribution[g,j,n] = S[g,j] × E[j,n]
  
  responsibility[g,j,n] = contribution[g,j,n] / Σ_j' contribution[g,j',n]
  
Interpretation: "What fraction of counts for gene g came from source j?"
```

### M-step: Update Exposures
```
For each source j:
  E[j,n] = Σ_g w[g] × k[g,n] × responsibility[g,j,n]
  
Interpretation: "Total weighted counts assigned to source j"

Intercept[n] = E[noise,n]  # Learned automatically
```

### Log-Likelihood (Monitoring)
```
ŷ = [S | 1/G] × [E; intercept] × (L / insert_size)
LL = Σ w[g] × [k·log(ŷ+ε) - ŷ - k·log(k) + k]

Guarantee: LL increases every iteration (or numerical precision)
```

---

## Optimization

**Single unified phase** (no null model needed!)

```
Initialize: E[c,n] = total_counts[n] / C
            intercept[n] = 0.05 × total_counts[n]

Repeat:
  1. E-step: Compute responsibilities for all sources
  2. M-step: Update E and intercept simultaneously
  3. Check: |ΔLL| < tolerance

Convergence: Typically 100-400 iterations
```

**With L1 regularization:**
```
After M-step: E ← max(E - λ₁, 0)  # Soft-thresholding
```

---

## Output

```
Exposures:   E[c,n]                              # Cell-equivalents
Intercept:   intercept[n]                        # Noise level
Proportions: P[c,n] = E[c,n] / Σ_c E[c,n]       # Fractions (sum to 1)
Quality:     intercept / (mean(E) + intercept)   # <0.1 = good fit
```

---

## Key Parameters

| Parameter | Default | Range | Purpose |
|-----------|---------|-------|---------|
| `l1_lambda` | 0 | 0-10 | Sparsity via soft-thresholding |
| `tolerance` | 10⁻⁶ | 10⁻⁸-10⁻⁴ | Convergence criterion (ΔLL) |
| `insert_size` | 500 | 150-500 | Fragment length correction |
| `max_iterations` | 1000 | 100-10000 | Safety limit |

**Recommendation:** Defaults work well. No learning rate tuning needed!

---

## Properties

- **Non-negativity:** Guaranteed by model structure (no transforms needed)
- **Monotonic improvement:** Log-likelihood never decreases
- **No hyperparameters:** Learning rate, momentum, warmup not needed
- **Natural intercept:** Noise competes with signatures automatically
- **Scale:** Set by total counts and signature normalization
- **Gene length correction:** Essential for RNA-seq (longer genes → more reads)
- **Performance:** 15-25 sec (typical: 18k genes, 21 types, 81 samples, 300 iters)

---

## Comparison to Gradient Descent

| Aspect | EM | Gradient Descent |
|--------|-----|------------------|
| **Convergence** | Guaranteed monotonic | Can oscillate/diverge |
| **Hyperparameters** | None (just tolerance) | Learning rate, warmup |
| **Iterations** | 100-400 | 1000-3000 |
| **Null model** | Not needed | Requires separate fit |
| **Interpretation** | Missing data framework | Optimization of loss |
| **Speed** | 15-25 sec | 20-30 sec |
| **Stability** | Very stable | Needs careful tuning |

---

## Intuition

**EM treats deconvolution as:**  
"For each read, which cell type (or noise) generated it?"

- **E-step:** Probabilistically assign reads to sources
- **M-step:** Count up assignments to estimate abundances

This "soft assignment" framework is more natural than gradient-based optimization for count data, leading to stable convergence without tuning.
