---
title: "Extra functions"
output: html_document
date: "2024-10-31"
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
rm(list=ls())
# Define dataset paths depending on your machine
if (grepl("^gizmo", Sys.info()["nodename"])) {
  ROOT_DIR1 <- "/fh/fast/furlan_s/experiments/MB_10X_5p/cds"
  ROOT_DIR2 <- "/fh/fast/furlan_s/grp/data/ddata/BM_data"
} else {
  ROOT_DIR1 <- "/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchinsonCancerCenter/Furlan_Lab - General/experiments/MB_10X_5p/cds"
  ROOT_DIR2 <- "/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchinsonCancerCenter/Furlan_Lab - General/datasets/Healthy_BM_greenleaf"
}

```


First, ensure you have the `devtools` R package installed, which allows you to install packages from GitHub. If `devtools` is installed, you can easily install viewmastR using the following command:

```{r, eval=F}
devtools::install_github("furlan-lab/viewmastR")
```

```{r}
# Load required packages
suppressPackageStartupMessages({
  library(viewmastR)
  library(Seurat)
  library(ggplot2)
  library(scCustomize)
  library(httpgd)
})


# Load query and reference datasets
seu <- readRDS(file.path(ROOT_DIR1, "240813_final_object.RDS"))
```

A clean Seurat object

```{r}
DimPlot(seu, group.by = "ground_truth")
```

Exporting R objects to scanpy is painful...  As an inefficient but effective work around we have written code that enables
you to export a Seurat object to the same three file format output by cellranger.  This makes import into scanpy a breeze.

The command creates a folder called 3file in the directory you provide.  By default the meta data, reductions, and variable features
are exported.

Currently, only Seurat objects are supported

```{r}
make3file(seu, dir = file.path(ROOT_DIR1))
list.files(file.path(ROOT_DIR1, "3file"))
```