---
title: "How to use viewmastR with saved models"
output: html_document
date: "2024-02-14"
always_allow_html: true
editor_options: 
  chunk_output_type: console
---


```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
rm(list=ls())
if(grepl("^gizmo", Sys.info()["nodename"])){
  ROOT_DIR1<-"/fh/fast/furlan_s/experiments/MB_10X_5p/cds"
  ROOT_DIR2<-"/fh/fast/furlan_s/grp/data/ddata/BM_data"
} else {
  ROOT_DIR1<-"/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchinsonCancerCenter/Furlan_Lab - General/experiments/MB_10X_5p/cds"
  ROOT_DIR2<-"/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchinsonCancerCenter/Furlan_Lab - General/datasets/Healthy_BM_greenleaf"
}
```


## Installing Rust

First you need to have an updated Rust installation. Go to this [site](https://www.rust-lang.org/tools/install) to learn how to install Rust.


## Installing viewmastR

You will need to have the devtools package installed...

```{r, eval=F}
devtools::install_github("furlan-lab/viewmastR")
```


## Running viewmastR

```{r, dpi=300, fig.height=4, fig.width = 6}
suppressPackageStartupMessages({
library(viewmastR)
library(Seurat)
library(ggplot2)
})

#query dataset
seu <- readRDS(file.path(ROOT_DIR1, "240813_final_object.RDS"))
#reference dataset
seur<-readRDS(file.path(ROOT_DIR2, "230329_rnaAugmented_seurat.RDS"))
vg <- get_selected_features(seu)


```

## Now you run viewmastR

The model path is specified using the 'dir' argument
```{r, dpi=300, fig.height=4, fig.width = 6, warning=F, message=F}
seu<-viewmastR(seu, seur, ref_celldata_col = "SFClassification", selected_features = vg, model_dir =  "/tmp/sc_local", max_epochs = 3, return_probs = T)
```

## A look at the predictions
```{r, dpi=300, fig.height=4, fig.width = 6}
DimPlot(seu, group.by = "viewmastR_pred", cols = seur@misc$colors)
```

## Run inference

We can use the function viewmastR_infer to run inference a saved model.  We will need to pass the same vector of variable features we used to initially create the model. We can use query_celldata_col to specify the name of the metadata column in the returned object.  An optional vector of labels can be provided.  Additionally, instead of returning the input object with predictions added, you may instead return the probabilities using the return_probs argument.
```{r}

seu_infer <- viewmastR_infer(seu, model_dir = "/tmp/sc_local", vg, 
                              labels = levels(factor(seur$SFClassification)),
                              return_probs = TRUE)

DimPlot(seu_infer, group.by = "viewmastR_inferred", cols = seur@misc$colors)


```


## Comparing training-time vs saved-model inference

When comparing predictions from viewmastR (training with simultaneous query inference) to viewmastR_infer (inference using a saved model), you may observe a small number of cells (~0.05%) with different classifications. These differences arise from minor numerical variations between the two computation paths.

Importantly, the disagreements occur exclusively for cells at classification boundaries—cases where the model assigns similar probabilities to multiple cell types. As shown below, the disagreeing cells typically have two or more competing classes with probabilities within a few percentage points of each other, indicating genuine ambiguity where either classification is reasonable.

```{r, dpi=300, fig.height=6, fig.width=10}

confusion_matrix(pred = factor(seu_infer$viewmastR_inferred), gt = factor(seu$viewmastR_pred), cols = seur@misc$colors)

disagree_idx <- which(seu$viewmastR_pred != seu_infer$viewmastR_inferred)
# Show what changed
disagree_df <- data.frame(
  cell = colnames(seu)[disagree_idx],
  viewmastR_pred = seu$viewmastR_pred[disagree_idx],
  viewmastR_inferred = seu_infer$viewmastR_inferred[disagree_idx]
)
print(disagree_df)


# For disagreeing cells, compare probabilities
if (length(disagree_idx) > 0) {
  probs_direct <- as.matrix(seu@meta.data[disagree_idx, prob_cols, drop = FALSE])
  probs_inferred <- as.matrix(seu_infer@meta.data[disagree_idx, prob_cols, drop = FALSE])
  
  # Calculate difference
  prob_diff <- probs_direct - probs_inferred
  
  # Show summary of absolute differences
  cat("\nAbsolute probability differences (disagreeing cells):\n")
  print(summary(as.vector(abs(prob_diff))))
  
  # For each disagreeing cell, show the two predictions and their probs
  disagree_df <- data.frame(
    cell = colnames(seu)[disagree_idx],
    pred_direct = seu$viewmastR_pred[disagree_idx],
    pred_inferred = seu_infer$viewmastR_inferred[disagree_idx],
    prob_direct = apply(probs_direct, 1, max),
    prob_inferred = apply(probs_inferred, 1, max),
    max_prob_diff = apply(abs(prob_diff), 1, max)
  )
  print(disagree_df)
  
  # Show the actual probabilities for the two competing classes
  for (i in seq_len(nrow(disagree_df))) {
    cell_idx <- disagree_idx[i]
    class_direct <- as.character(disagree_df$pred_direct[i])
    class_inferred <- as.character(disagree_df$pred_inferred[i])
    
    col_direct <- paste0("prob_", class_direct)
    col_inferred <- paste0("prob_", class_inferred)
    
    cat(sprintf("\nCell %s:\n", disagree_df$cell[i]))
    cat(sprintf("  Direct:   %s = %.6f, %s = %.6f\n", 
                class_direct, seu@meta.data[cell_idx, col_direct],
                class_inferred, seu@meta.data[cell_idx, col_inferred]))
    cat(sprintf("  Inferred: %s = %.6f, %s = %.6f\n", 
                class_direct, seu_infer@meta.data[cell_idx, col_direct],
                class_inferred, seu_infer@meta.data[cell_idx, col_inferred]))
  }
}
```

## Appendix
```{r Appendix}
sessionInfo()
getwd()
```
